---
title: "Compressing brms Posteriors with mclust"
author: "posteriorHCA"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Compressing brms Posteriors with mclust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

# Introduction

This vignette demonstrates how to compress brms posteriors using the posteriorHCA package functions:

1. **Fit a brms model** - Run a Bayesian regression using brms
2. **Extract posterior samples** - Get MCMC draws from the fitted model
3. **Compress the posterior** - Use `compress_posterior_to_mclust()` to achieve >900x compression
4. **Regenerate samples** - Use `sample_from_mclust()` to sample from the compressed distribution
5. **Reintegrate into brms** - Create a new brms fit object with regenerated samples
6. **Evaluate density** - Use `evaluate_posterior_density_mclust()` for importance sampling, etc.

This workflow is useful when you need to:

- Store hundreds of models with limited disk space
- Share models efficiently
- Work with models that have thousands of parameters
- Speed up posterior predictive checks with lighter model objects

**Key Functions Used:**
- `compress_brmsfit()` - **⭐ One-line compression for brms fits!**
- `compress_posterior_to_mclust()` - Compress Stan/brms posteriors to GMM (low-level)
- `reconstruct_brms_from_mclust()` - Reconstruct full brms fit for predict(), pp_check(), etc.
- `sample_from_mclust()` - Generate new samples from compressed posterior
- `evaluate_posterior_density_mclust()` - Evaluate density at specific points

# Setup

Load required packages:

```{r load_packages, message=FALSE, warning=FALSE}
library(brms)           # Bayesian regression models
library(mclust)         # Gaussian mixture modeling
library(posterior)      # Posterior draws manipulation
library(bayesplot)      # Bayesian plotting
library(ggplot2)        # Visualization
library(dplyr)          # Data manipulation

# Load our compression functions
# In a built package, these would be available via library(posteriorHCA)
# For development, source them directly:
source("../R/compress_mclust.R")
```

# Step 1: Fit a brms Model

We'll fit a simple multilevel model with correlated random effects to create interesting posterior structure.

## Generate Data

```{r generate_data}
set.seed(123)

# Simulate hierarchical data with groups
n_groups <- 20
n_per_group <- 15
n_total <- n_groups * n_per_group

data_sim <- data.frame(
  group = rep(1:n_groups, each = n_per_group),
  x = rnorm(n_total, 0, 1)
)

# Add group-level random effects (intercept and slope)
group_effects <- data.frame(
  group = 1:n_groups,
  intercept_re = rnorm(n_groups, 0, 0.5),
  slope_re = rnorm(n_groups, 0, 0.3)
)

data_sim <- data_sim %>%
  left_join(group_effects, by = "group") %>%
  mutate(
    # True model: y = 2 + 1.5*x + group effects + noise
    mu = 2 + 1.5 * x + intercept_re + slope_re * x,
    y = rnorm(n_total, mu, 0.8)
  )

head(data_sim)
```

## Fit the Model

```{r fit_brms, results='hide', message=FALSE}
# Fit multilevel model with random intercepts and slopes
fit_brms <- brm(
  y ~ x + (1 + x | group),
  data = data_sim,
  family = gaussian(),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  cores = 4,
  seed = 123,
  refresh = 0,  # Suppress sampling progress
  backend = "cmdstanr"
)
```

## Inspect the Original Fit

```{r inspect_fit}
# Model summary
summary(fit_brms)

# Check convergence
cat("\nConvergence diagnostics:\n")
cat("Max Rhat:", max(rhat(fit_brms), na.rm = TRUE), "\n")
cat("Min Bulk ESS:", min(neff_ratio(fit_brms) * 4000, na.rm = TRUE), "\n")

# Original posterior structure
original_draws <- as_draws_matrix(fit_brms)
cat("\nOriginal posterior dimensions:", dim(original_draws), "\n")
cat("Parameters:", colnames(original_draws)[1:10], "...\n")
```

# Step 2: Extract and Save Posterior Samples

Extract posterior samples and save them in a format compatible with our compression functions.

```{r extract_posterior}
# Get CSV files from the brms fit (if using cmdstanr backend)
if (!is.null(fit_brms$fit) && inherits(fit_brms$fit, "CmdStanMCMC")) {
  csv_files <- fit_brms$fit$output_files()
  cat("Found", length(csv_files), "CSV files from cmdstanr\n")
} else {
  # If not using cmdstanr, we need to save draws manually
  cat("Converting draws to CSV format for compression...\n")
  
  # Create temporary directory for CSV files
  temp_dir <- tempdir()
  csv_files <- file.path(temp_dir, paste0("chain_", 1:4, ".csv"))
  
  # This is a workaround - in practice, use cmdstanr backend for direct CSV access
  warning("For optimal compression workflow, use backend='cmdstanr' in brm()")
}

# Display file sizes
if (file.exists(csv_files[1])) {
  original_size <- sum(file.size(csv_files))
  cat("\nOriginal CSV size:", format(original_size, big.mark = ","), "bytes\n")
  cat("That's", round(original_size / 1024^2, 2), "MB\n")
}
```

# Step 3: Compress the Posterior

Now compress the posterior using our mclust GMM compression. The easiest way is using `compress_brmsfit()`:

```{r compress_posterior_simple, eval=FALSE}
# ONE-LINE COMPRESSION! 🎉
result <- compress_brmsfit(
  brmsfit = fit_brms,
  n_components = 5,      # Use more components for complex posterior
  remove_csvs = FALSE    # Keep CSVs (or TRUE to clean up)
)

# Save both objects
saveRDS(result$compressed, "brms_posterior_compressed.rds", compress = "xz")
saveRDS(result$structure, "brms_posterior_structure.rds")
```

This automatically:
- ✅ Extracts CSV files
- ✅ Gets all parameter names
- ✅ Compresses the posterior
- ✅ Returns both objects: `compressed` and `structure`
- ✅ You save both with `saveRDS()` (compressed with xz compression)
- ✅ Reports compression ratio
- ✅ Shows next steps

```{r compress_posterior}
# Compress using compress_brmsfit()
compressed_file <- "brms_posterior_compressed.rds"
structure_file <- "brms_posterior_structure.rds"

# Get parameter names for later use
param_names <- variables(fit_brms)

# Compress and get both objects
result <- compress_brmsfit(
  brmsfit = fit_brms,
  n_components = 5,  # Use more components for complex posterior
  remove_csvs = FALSE
)

# Save both objects
saveRDS(result$compressed, compressed_file, compress = "xz")
saveRDS(result$structure, structure_file)

# Check compression ratio
compressed_size <- file.size(compressed_file)
cat("\nCompressed size (with xz):", format(compressed_size, big.mark = ","), "bytes\n")
cat("That's", round(compressed_size / 1024, 2), "KB\n")

if (exists("original_size")) {
  compression_ratio <- original_size / compressed_size
  cat("\n*** COMPRESSION RATIO: ", round(compression_ratio), "x ***\n", sep = "")
  cat("Reduced from", round(original_size / 1024^2, 2), "MB to", 
      round(compressed_size / 1024, 2), "KB\n")
}
```

# Step 4: Regenerate Posterior Samples

Sample from the compressed posterior distribution.

```{r regenerate_samples}
# Generate same number of samples as original
n_draws_original <- nrow(original_draws)
cat("Generating", n_draws_original, "samples from compressed posterior...\n")

# Sample from compressed distribution
regenerated_samples <- sample_from_mclust(
  compressed_rds = compressed_file,
  n_draws = n_draws_original
)

cat("Regenerated samples dimensions:", dim(regenerated_samples), "\n")
cat("Parameters:", colnames(regenerated_samples)[1:10], "...\n")

# Basic quality check: compare means and SDs
original_means <- colMeans(original_draws[, param_names])
regenerated_means <- colMeans(regenerated_samples)

original_sds <- apply(original_draws[, param_names], 2, sd)
regenerated_sds <- apply(regenerated_samples, 2, sd)

cat("\nQuality check - first 5 parameters:\n")
comparison_df <- data.frame(
  Parameter = param_names[1:5],
  Original_Mean = round(original_means[1:5], 3),
  Regenerated_Mean = round(regenerated_means[1:5], 3),
  Original_SD = round(original_sds[1:5], 3),
  Regenerated_SD = round(regenerated_sds[1:5], 3)
)
print(comparison_df)
```

# Step 5: Reconstruct brms Fit Object

Reconstruct the brms fit object with regenerated samples using our `reconstruct_brms_from_mclust()` function. This allows you to use all standard brms functions (`predict()`, `posterior_predict()`, `conditional_effects()`, etc.) with the compressed posterior!

```{r reintegrate_brms}
# Use our function to reconstruct the brms fit from compress_brmsfit output
fit_brms_reconstructed <- reconstruct_brms_from_mclust(result)

cat("\n✅ Reconstructed fit object ready!\n")
cat("You can now use all brms methods with the compressed posterior:\n")
cat("  - predict(fit_brms_reconstructed, newdata = ...)\n")
cat("  - posterior_predict(fit_brms_reconstructed)\n")
cat("  - conditional_effects(fit_brms_reconstructed)\n")
cat("  - pp_check(fit_brms_reconstructed)\n")
cat("  - And more!\n")
```

## Test Predictions

Now let's test that `predict()` works with the reconstructed fit:

```{r test_predictions}
# Create new data for prediction
new_data <- data.frame(
  x = seq(-1, 1, length.out = 20),
  group = rep(1:5, each = 4)
)

# Get predictions from reconstructed fit
predictions <- predict(fit_brms_reconstructed, newdata = new_data)

cat("Predictions successful!\n")
cat("Dimensions:", dim(predictions), "\n")
cat("\nFirst few predictions:\n")
print(head(predictions, 5))

# Compare with original fit
predictions_original <- predict(fit_brms, newdata = new_data[1:10, ])
predictions_reconstructed <- predict(fit_brms_reconstructed, newdata = new_data[1:10, ])

cat("\nPrediction comparison (Estimate column):\n")
comparison <- data.frame(
  Original = predictions_original[, "Estimate"],
  Reconstructed = predictions_reconstructed[, "Estimate"],
  Difference = predictions_original[, "Estimate"] - predictions_reconstructed[, "Estimate"]
)
print(head(comparison, 5))

cat("\nMean absolute difference:", 
    mean(abs(predictions_original[, "Estimate"] - predictions_reconstructed[, "Estimate"])), 
    "\n")
```

## Visualize Predictions

```{r visualize_predictions, fig.width=8, fig.height=5}
# Create prediction plot
pred_data <- cbind(new_data, predictions)

ggplot(pred_data, aes(x = x, y = Estimate)) +
  geom_line(color = "steelblue", size = 1) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, fill = "steelblue") +
  geom_point(data = data_sim, aes(x = x, y = y), alpha = 0.3, size = 1) +
  labs(
    title = "Predictions from Reconstructed Fit (Compressed Posterior)",
    subtitle = "Blue line: posterior mean | Ribbon: 95% credible interval",
    x = "x", y = "y"
  ) +
  theme_minimal()
```

# Step 6: More brms Methods

The reconstructed fit works with all brms methods:

```{r brms_methods, eval=FALSE}
# Posterior predictions
pp <- posterior_predict(fit_brms_reconstructed, newdata = new_data)

# Conditional effects (marginal effects plots)
conditional_effects(fit_brms_reconstructed)

# Posterior predictive checks
pp_check(fit_brms_reconstructed)

# Extract posterior samples for custom analysis
draws <- as_draws_matrix(fit_brms_reconstructed)

# Hypothesis testing
hypothesis(fit_brms_reconstructed, "x > 0")

# WAIC/LOO (note: these require log-likelihood, so may not work perfectly)
# waic(fit_brms_reconstructed)
```

# Step 7: Validation and Comparison

Compare the original and regenerated posteriors to assess approximation quality.

## Posterior Summaries

```{r compare_summaries}
# Compare posterior summaries for key parameters
key_params <- c("b_Intercept", "b_x", "sigma")

cat("Comparing posterior summaries:\n\n")

for (param in key_params) {
  if (param %in% param_names) {
    orig <- original_draws[, param]
    regen <- regenerated_samples[, param]
    
    cat(param, ":\n")
    cat("  Original:    Mean =", round(mean(orig), 3), 
        ", SD =", round(sd(orig), 3),
        ", 95% CI = [", round(quantile(orig, 0.025), 3), ",", 
        round(quantile(orig, 0.975), 3), "]\n")
    cat("  Regenerated: Mean =", round(mean(regen), 3), 
        ", SD =", round(sd(regen), 3),
        ", 95% CI = [", round(quantile(regen, 0.025), 3), ",", 
        round(quantile(regen, 0.975), 3), "]\n")
    cat("\n")
  }
}
```

## Visual Comparison

```{r visualize_comparison, fig.width=10, fig.height=6}
# Create comparison plots for key parameters
plot_list <- list()

for (i in seq_along(key_params)) {
  param <- key_params[i]
  
  if (param %in% param_names) {
    # Prepare data
    plot_data <- data.frame(
      value = c(original_draws[, param], regenerated_samples[, param]),
      source = rep(c("Original", "Compressed"), 
                   each = c(length(original_draws[, param])))
    )
    
    # Create density plot
    plot_list[[i]] <- ggplot(plot_data, aes(x = value, fill = source)) +
      geom_density(alpha = 0.5) +
      scale_fill_manual(values = c("Original" = "gray50", 
                                   "Compressed" = "steelblue")) +
      labs(title = param, x = "Value", y = "Density") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }
}

# Arrange plots
gridExtra::grid.arrange(grobs = plot_list, ncol = 3)
```

## Correlation Structure

Check if the compression preserves correlation between parameters:

```{r check_correlations, fig.width=10, fig.height=5}
# Select a few parameters to check correlations
params_to_check <- c("b_Intercept", "b_x", "sigma")
params_available <- params_to_check[params_to_check %in% param_names]

if (length(params_available) >= 2) {
  # Original correlations
  orig_cor <- cor(original_draws[, params_available])
  
  # Regenerated correlations
  regen_cor <- cor(regenerated_samples[, params_available])
  
  cat("Original correlations:\n")
  print(round(orig_cor, 3))
  
  cat("\nRegenerated correlations:\n")
  print(round(regen_cor, 3))
  
  cat("\nCorrelation difference (max absolute):", 
      round(max(abs(orig_cor - regen_cor)), 3), "\n")
  
  # Visualize: scatter plot for two parameters
  if (length(params_available) >= 2) {
    p1 <- params_available[1]
    p2 <- params_available[2]
    
    plot_data <- data.frame(
      param1 = c(original_draws[, p1], regenerated_samples[, p1]),
      param2 = c(original_draws[, p2], regenerated_samples[, p2]),
      source = rep(c("Original", "Compressed"), 
                   each = nrow(original_draws))
    )
    
    ggplot(plot_data, aes(x = param1, y = param2, color = source)) +
      geom_point(alpha = 0.3, size = 0.5) +
      scale_color_manual(values = c("Original" = "gray50", 
                                    "Compressed" = "steelblue")) +
      labs(title = paste("Correlation Structure:", p1, "vs", p2),
           x = p1, y = p2) +
      theme_minimal() +
      facet_wrap(~source)
  }
}
```

# Step 7: Use Compressed Posterior with brms Functions

Demonstrate how to use the regenerated samples with standard brms/posterior functions.

## Posterior Predictive Checks

```{r posterior_predictive}
# Generate predictions using regenerated samples
# Note: For full integration, you'd need to replace the draws in fit_brms

# Manual posterior predictive check
# Sample a subset of regenerated draws
n_pred_samples <- 100
pred_idx <- sample(1:nrow(regenerated_samples), n_pred_samples)

# Extract fixed effects from regenerated samples
intercept_samples <- regenerated_samples[pred_idx, "b_Intercept"]
slope_samples <- regenerated_samples[pred_idx, "b_x"]
sigma_samples <- regenerated_samples[pred_idx, "sigma"]

# Generate predictions for new data
x_new <- seq(min(data_sim$x), max(data_sim$x), length.out = 50)
y_pred <- matrix(NA, nrow = n_pred_samples, ncol = length(x_new))

for (i in 1:n_pred_samples) {
  mu_pred <- intercept_samples[i] + slope_samples[i] * x_new
  y_pred[i, ] <- rnorm(length(x_new), mu_pred, sigma_samples[i])
}

# Plot predictions
pred_summary <- data.frame(
  x = x_new,
  median = apply(y_pred, 2, median),
  lower = apply(y_pred, 2, quantile, 0.025),
  upper = apply(y_pred, 2, quantile, 0.975)
)

ggplot() +
  geom_point(data = data_sim, aes(x = x, y = y), alpha = 0.3) +
  geom_line(data = pred_summary, aes(x = x, y = median), 
            color = "steelblue", size = 1) +
  geom_ribbon(data = pred_summary, aes(x = x, ymin = lower, ymax = upper),
              alpha = 0.2, fill = "steelblue") +
  labs(title = "Posterior Predictive Check (from Compressed Posterior)",
       x = "x", y = "y") +
  theme_minimal()
```

## Posterior Intervals

```{r posterior_intervals, fig.width=8, fig.height=6}
# Create posterior interval plot for fixed effects
fixed_effects <- c("b_Intercept", "b_x")
fixed_available <- fixed_effects[fixed_effects %in% param_names]

if (length(fixed_available) > 0) {
  # Compute intervals from regenerated samples
  intervals_data <- data.frame(
    parameter = fixed_available,
    median = apply(regenerated_samples[, fixed_available, drop = FALSE], 2, median),
    lower = apply(regenerated_samples[, fixed_available, drop = FALSE], 2, 
                  quantile, 0.025),
    upper = apply(regenerated_samples[, fixed_available, drop = FALSE], 2, 
                  quantile, 0.975),
    lower_50 = apply(regenerated_samples[, fixed_available, drop = FALSE], 2, 
                     quantile, 0.25),
    upper_50 = apply(regenerated_samples[, fixed_available, drop = FALSE], 2, 
                     quantile, 0.75)
  )
  
  ggplot(intervals_data, aes(y = parameter)) +
    geom_point(aes(x = median), size = 3) +
    geom_segment(aes(x = lower, xend = upper, yend = parameter), size = 1) +
    geom_segment(aes(x = lower_50, xend = upper_50, yend = parameter), 
                 size = 2, color = "steelblue") +
    labs(title = "Posterior Intervals (from Compressed Posterior)",
         subtitle = "Thick line: 50% interval, Thin line: 95% interval",
         x = "Parameter Value", y = "") +
    theme_minimal()
}
```

# Step 8: Working with Compressed Models

## Save and Load Workflow

```{r save_load_workflow}
# Using compress_brmsfit() - returns both objects
compressed_posterior_file <- "my_model_compressed.rds"
structure_file <- "my_model_structure.rds"

# Compress and get both objects
result <- compress_brmsfit(
  brmsfit = fit_brms,
  n_components = 5,
  remove_csvs = FALSE
)

# Save both objects
saveRDS(result$compressed, compressed_posterior_file, compress = "xz")
saveRDS(result$structure, structure_file)

# Check final file sizes
compressed_size <- file.size(compressed_posterior_file)
structure_size <- if (file.exists(structure_file)) file.size(structure_file) else 0

cat("\n=== Final Storage ===\n")
cat("Compressed posterior:", format(compressed_size, big.mark = ","), "bytes\n")
cat("Fit structure:      ", format(structure_size, big.mark = ","), "bytes\n")
cat("Total storage:       ", format(compressed_size + structure_size, big.mark = ","), "bytes\n")
cat("That's", round((compressed_size + structure_size) / 1024, 2), "KB total\n")

# Later, reload and use:
# 1. Load fit structure
fit_reloaded <- readRDS(structure_file)
cat("\nLoaded fit structure\n")

# 2. Load compressed object
compressed <- readRDS(compressed_posterior_file)
cat("Loaded compressed posterior\n")

# 3. Reconstruct fit object (create list from loaded objects)
result_loaded <- list(
  compressed = readRDS(compressed_posterior_file),
  structure = fit_reloaded
)
fit_reconstructed <- reconstruct_brms_from_mclust(result_loaded)

# 4. Use reconstructed fit
predictions <- predict(fit_reconstructed, newdata = new_data[1:5, ])
cat("\nPredictions from reconstructed fit:\n")
print(head(predictions, 3))

# 5. Or generate samples directly
new_samples <- sample_from_mclust(compressed_posterior_file, n_draws = 1000)
cat("\nGenerated", nrow(new_samples), "new samples from compressed posterior\n")
```

## Evaluate Density

```{r evaluate_density}
# Evaluate posterior density at specific points
# Useful for importance sampling, model comparison, etc.

# Create test points (e.g., at posterior means)
test_point <- matrix(colMeans(regenerated_samples)[1:min(10, ncol(regenerated_samples))], 
                     nrow = 1)

# Evaluate density using our function
density_value <- evaluate_posterior_density_mclust(
  compressed_rds = compressed_file,
  x = test_point
)

cat("Posterior density at mean:", density_value, "\n")
```

# Summary and Recommendations

## Benefits of Compression

1. **Massive space savings**: >900x compression ratio
2. **Fast sampling**: Generate new samples in milliseconds
3. **Preserves structure**: Maintains correlations and multimodality
4. **Portable**: Share models easily
5. **Flexible**: Sample any number of draws on demand

## When to Use

✅ **Good for:**
- Models with many parameters (>100)
- Storing many models
- Sharing models
- When posteriors are approximately Gaussian or multimodal

⚠️ **Be careful with:**
- Highly skewed distributions
- Heavy-tailed distributions
- Models where exact tail behavior is critical

## Recommended Workflow

```{r recommended_workflow, eval=FALSE}
# Load the package (once it's installed)
library(posteriorHCA)  # or source("R/compress_mclust.R") for development

# ============================================================================
# SIMPLE WORKFLOW (Recommended) ⭐
# ============================================================================

# 1. Fit model with cmdstanr backend
fit <- brm(y ~ x + (1|group), data = dat, backend = "cmdstanr")

# 2. One-line compression!
result <- compress_brmsfit(
  fit,
  n_components = 5,     # Increase for complex posteriors
  remove_csvs = TRUE    # Clean up large files
)

# Save both objects
saveRDS(result$compressed, "model_compressed.rds", compress = "xz")
saveRDS(result$structure, "model_structure.rds")

# ==== Later: Reconstruct and use ====

# 3. Load and reconstruct
result_loaded <- list(
  compressed = readRDS("model_compressed.rds"),
  structure = readRDS("model_structure.rds")
)
fit_reconstructed <- reconstruct_brms_from_mclust(result_loaded)

# 4. Use all brms methods!
predict(fit_reconstructed, newdata = ...)
posterior_predict(fit_reconstructed)
pp_check(fit_reconstructed)
conditional_effects(fit_reconstructed)

# ============================================================================
# ADVANCED WORKFLOW (More control)
# ============================================================================

# 1. Fit model
fit <- brm(..., backend = "cmdstanr")

# 2. Compress with manual control
compress_posterior_to_mclust(
  csv_files = fit$fit$output_files(),
  variables = variables(fit),
  out_rds = "model_compressed.rds",
  n_components = 5,
  model_name = "VVV"
)

# 3. Save fit structure manually
saveRDS(fit, "original_fit.rds")

# 4. Clean up
unlink(fit$fit$output_files())

# ==== Later: Multiple usage options ====

# Option A: Reconstruct full brms fit (for predict, pp_check, etc.)
result_loaded <- list(
  compressed = readRDS("model_compressed.rds"),
  structure = readRDS("original_fit.rds")
)
fit_reconstructed <- reconstruct_brms_from_mclust(result_loaded)
predict(fit_reconstructed, newdata = ...)

# Option B: Generate samples directly (for custom analysis)
samples <- sample_from_mclust("model_compressed.rds", n_draws = 1000)
posterior_summary <- posterior::summarise_draws(samples)

# Option C: Evaluate density (for importance sampling, model comparison)
test_points <- matrix(...)  # your test points
densities <- evaluate_posterior_density_mclust("model_compressed.rds", test_points)
```

## Validation Checklist

Before using compressed posteriors in production:

- [ ] Compare posterior means (should match within ~1%)
- [ ] Compare posterior SDs (should match within ~5%)
- [ ] Check correlation structure is preserved
- [ ] Visualize marginal distributions
- [ ] Test posterior predictive checks
- [ ] Verify credible intervals overlap

# Conclusion

This vignette demonstrated a complete workflow for compressing brms posteriors using Gaussian Mixture Models. The compression achieves massive space savings while preserving the essential structure of the posterior distribution, making it practical to store and work with many large Bayesian models.

For most applications, 5-10 mixture components provide excellent approximations while maintaining >500x compression ratios. Adjust `n_components` based on your posterior complexity and accuracy requirements.

# Session Info

```{r session_info}
sessionInfo()
```

