---
title: "Funnel Posterior Compression with mclust"
author: "posteriorHCA"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 7,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

This document demonstrates how to compress and regenerate a **funnel posterior** using Gaussian Mixture Models (GMM) via the `mclust` package. The funnel posterior is a classic challenging distribution in Bayesian statistics, often used to test MCMC samplers.

## What is a Funnel Posterior?

The funnel posterior is a bivariate distribution with a characteristic funnel shape:

- $x \sim \text{Normal}(0, 3)$
- $y \mid x \sim \text{Normal}(0, \exp(x/2))$

This creates a distribution where:
- The variance of $y$ depends exponentially on $x$
- The distribution is narrow at the top (large $x$) and wide at the bottom (small $x$)
- It's challenging for MCMC samplers due to the varying scales

## Objectives

1. Generate samples from a funnel posterior
2. Visualize the original posterior distribution
3. Compress the posterior using `compress_posterior_to_mclust`
4. Regenerate samples using `sample_from_mclust`
5. Compare original vs. regenerated distributions
6. Evaluate compression quality and performance

# Setup

```{r load_packages}
if (!requireNamespace("posteriorHCA", quietly = TRUE)) {
  if (requireNamespace("pkgload", quietly = TRUE)) {
    pkgload::load_all("..", quiet = TRUE)
  } else {
    stop("posteriorHCA is not installed and pkgload is unavailable. Install the package or install.packages('pkgload').")
  }
}
library(posteriorHCA)  # Our compression functions
library(cmdstanr)      # CmdStan utilities for CSV handling
library(mclust)        # Gaussian mixture modeling
library(ggplot2)       # Visualization
library(dplyr)         # Data manipulation
library(viridis)       # Color scales
library(gridExtra)     # Multiple plots

# Set seed for reproducibility
set.seed(123)
```

# Step 1: Sample the Funnel Posterior with Stan

Rather than simulating by hand, we'll fit a tiny Stan model that directly samples from the funnel posterior. This automatically gives us CmdStan CSV files that plug into `compress_posterior_to_mclust()`.

```{r generate_funnel_samples}
# Stan program for funnel posterior
funnel_stan_code <- "
parameters {
  real x;
  real y;
}
model {
  x ~ normal(0, 3);
  y ~ normal(0, exp(x / 2));
}
"

# Write Stan model to temporary file
temp_dir <- tempdir()
stan_file <- file.path(temp_dir, "funnel_model.stan")
writeLines(funnel_stan_code, stan_file)

# Compile model (cached if file unchanged)
funnel_mod <- cmdstan_model(stan_file)

# Sample from the model
n_chains <- 4
iter_warmup <- 1000
iter_sampling <- 2000
funnel_fit <- funnel_mod$sample(
  chains = n_chains,
  iter_warmup = iter_warmup,
  iter_sampling = iter_sampling,
  refresh = 500,
  seed = 123
)

# Extract draws matrix for analysis and plotting
funnel_samples <- funnel_fit$draws(
  variables = c("x", "y"),
  format = "matrix"
)

n_samples <- nrow(funnel_samples)
funnel_df <- data.frame(
  x = funnel_samples[, "x"],
  y = funnel_samples[, "y"],
  source = "original"
)

csv_files <- funnel_fit$output_files()

cat("Generated", n_samples, "posterior draws via CmdStan\n")
cat("CSV files stored at:\n")
cat(paste("  -", csv_files), sep = "\n")
```

## Visualize Original Funnel Posterior

```{r visualize_original}
# Create scatter plot with density
p1 <- ggplot(funnel_df, aes(x = x, y = y)) +
  geom_point(alpha = 0.3, size = 0.5, color = "steelblue") +
  stat_density_2d(aes(fill = after_stat(level)), geom = "polygon", alpha = 0.5) +
  scale_fill_viridis_c(name = "Density") +
  labs(
    title = "Original Funnel Posterior",
    subtitle = paste("n =", n_samples, "samples"),
    x = "x ~ Normal(0, 3)",
    y = "y | x ~ Normal(0, exp(x/2))"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# Marginal distributions
p2 <- ggplot(funnel_df, aes(x = x)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Marginal: x", x = "x", y = "Density") +
  theme_minimal()

p3 <- ggplot(funnel_df, aes(x = y)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(title = "Marginal: y", x = "y", y = "Density") +
  theme_minimal()

# Combine plots
grid.arrange(p1, arrangeGrob(p2, p3, ncol = 2), nrow = 2, heights = c(2, 1))
```

# Step 2: Compress the Funnel Posterior

Now we'll compress the posterior using `compress_posterior_to_mclust`. We'll try different numbers of components to see how well the GMM captures the funnel shape.

```{r compress_funnel}
# Compress with different numbers of components
n_components_options <- c(3, 5, 10)

compressed_results <- list()
compression_stats <- data.frame()

for (n_comp in n_components_options) {
  cat("\n=== Compressing with", n_comp, "components ===\n")
  
  # Output file for compressed posterior
  out_file <- file.path(temp_dir, paste0("funnel_compressed_", n_comp, ".rds"))
  
  # Compress
  start_time <- Sys.time()
  compressed <- compress_posterior_to_mclust(
    csv_files = csv_files,
    variables = c("x", "y"),
    out_rds = out_file,
    n_components = n_comp,
    model_name = "VVV"  # Variable volume, shape, orientation
  )
  compression_time <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))
  
  # Calculate compression ratio
  original_size <- sum(file.size(csv_files))
  compressed_size <- file.size(out_file)
  compression_ratio <- original_size / compressed_size
  
  # Store results
  compressed_results[[as.character(n_comp)]] <- list(
    compressed = compressed,
    file = out_file,
    n_components = n_comp
  )
  
  # Store stats
  compression_stats <- rbind(compression_stats, data.frame(
    n_components = n_comp,
    compression_time = compression_time,
    original_size_mb = original_size / 1024^2,
    compressed_size_kb = compressed_size / 1024,
    compression_ratio = compression_ratio,
    loglik = compressed$loglik,
    bic = compressed$bic
  ))
  
  cat("Compression time:", round(compression_time, 2), "seconds\n")
  cat("Compression ratio:", round(compression_ratio, 1), "x\n")
  cat("BIC:", round(compressed$bic, 2), "\n")
}

# Display compression statistics
print(compression_stats)
```

## Select Best Model

We'll use the model with the best BIC (lower is better, but we also want enough components to capture the funnel shape).

```{r select_best_model}
# Select model (we'll use 10 components for better capture of funnel shape)
selected_n_comp <- 10
selected_compressed <- compressed_results[[as.character(selected_n_comp)]]

cat("Selected model with", selected_n_comp, "components\n")
cat("BIC:", round(selected_compressed$compressed$bic, 2), "\n")
cat("Log-likelihood:", round(selected_compressed$compressed$loglik, 2), "\n")
```

# Step 4: Regenerate Samples from Compressed Posterior

Now we'll generate new samples from the compressed GMM using `sample_from_mclust`.

```{r regenerate_samples}
# Regenerate samples
n_regenerated <- n_samples

cat("Regenerating", n_regenerated, "samples from compressed posterior...\n")

start_time <- Sys.time()
regenerated_samples <- sample_from_mclust(
  compressed_rds = selected_compressed$file,
  n_draws = n_regenerated
)
regeneration_time <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))

cat("Regeneration time:", round(regeneration_time * 1000, 2), "milliseconds\n")
cat("Samples per second:", round(n_regenerated / regeneration_time, 0), "\n")

# Convert to data frame
regenerated_df <- data.frame(
  x = regenerated_samples[, "x"],
  y = regenerated_samples[, "y"],
  source = "regenerated"
)

# Combine original and regenerated for comparison
comparison_df <- rbind(
  funnel_df,
  regenerated_df
)
```

# Step 5: Compare Original vs. Regenerated

## Side-by-Side Visualization

```{r compare_visualizations}
# Original
p_orig <- ggplot(funnel_df, aes(x = x, y = y)) +
  geom_point(alpha = 0.3, size = 0.5, color = "steelblue") +
  stat_density_2d(aes(fill = after_stat(level)), geom = "polygon", alpha = 0.5) +
  scale_fill_viridis_c(name = "Density") +
  labs(
    title = "Original Funnel Posterior",
    x = "x",
    y = "y"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Regenerated
p_regen <- ggplot(regenerated_df, aes(x = x, y = y)) +
  geom_point(alpha = 0.3, size = 0.5, color = "darkgreen") +
  stat_density_2d(aes(fill = after_stat(level)), geom = "polygon", alpha = 0.5) +
  scale_fill_viridis_c(name = "Density") +
  labs(
    title = "Regenerated from GMM",
    x = "x",
    y = "y"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Overlay comparison
p_overlay <- ggplot(comparison_df, aes(x = x, y = y, color = source)) +
  geom_point(alpha = 0.2, size = 0.5) +
  stat_density_2d(aes(fill = source), alpha = 0.3, geom = "polygon") +
  scale_color_manual(values = c("original" = "steelblue", "regenerated" = "darkgreen")) +
  scale_fill_manual(values = c("original" = "steelblue", "regenerated" = "darkgreen")) +
  labs(
    title = "Overlay: Original (blue) vs. Regenerated (green)",
    x = "x",
    y = "y"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Arrange plots
grid.arrange(p_orig, p_regen, p_overlay, ncol = 3)
```

## Marginal Distribution Comparison

```{r compare_marginals}
# Marginal x
p_x <- ggplot(comparison_df, aes(x = x, fill = source)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("original" = "steelblue", "regenerated" = "darkgreen")) +
  labs(title = "Marginal: x", x = "x", y = "Density", fill = "Source") +
  theme_minimal()

# Marginal y
p_y <- ggplot(comparison_df, aes(x = y, fill = source)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("original" = "steelblue", "regenerated" = "darkgreen")) +
  labs(title = "Marginal: y", x = "y", y = "Density", fill = "Source") +
  theme_minimal()

grid.arrange(p_x, p_y, ncol = 2)
```

## Statistical Comparison

```{r statistical_comparison}
# Calculate summary statistics
stats_comparison <- comparison_df %>%
  group_by(source) %>%
  summarise(
    mean_x = mean(x),
    sd_x = sd(x),
    mean_y = mean(y),
    sd_y = sd(y),
    cor_xy = cor(x, y),
    .groups = "drop"
  )

print(stats_comparison)

# Calculate differences
diff_stats <- data.frame(
  statistic = c("mean_x", "sd_x", "mean_y", "sd_y", "cor_xy"),
  original = as.numeric(stats_comparison[stats_comparison$source == "original", -1]),
  regenerated = as.numeric(stats_comparison[stats_comparison$source == "regenerated", -1])
) %>%
  mutate(
    difference = regenerated - original,
    relative_error = abs(difference / original) * 100
  )

print(diff_stats)
```

# Step 6: Evaluate Density Approximation

We can evaluate how well the GMM approximates the true density at specific points.

```{r evaluate_density}
# Select some test points
test_points <- data.frame(
  x = c(-3, 0, 3, -2, 0, 2),
  y = c(0, 0, 0, 1, 1, 1)
)

# Evaluate density using compressed posterior
densities <- evaluate_posterior_density_mclust(
  compressed_rds = selected_compressed$file,
  x = as.matrix(test_points)
)

test_points$density <- densities

cat("Density evaluation at test points:\n")
print(test_points)

# Visualize density surface
p_density <- ggplot(test_points, aes(x = x, y = y, size = density, color = density)) +
  geom_point() +
  scale_size_continuous(range = c(3, 10), name = "Density") +
  scale_color_viridis_c(name = "Density") +
  labs(
    title = "Density at Test Points",
    x = "x",
    y = "y"
  ) +
  theme_minimal()

print(p_density)
```

# Step 7: Component Visualization

Let's visualize the GMM components to see how they capture the funnel shape.

```{r visualize_components}
# Extract component information
compressed <- selected_compressed$compressed
n_comp <- compressed$n_components

# Create plot showing component centers and ellipses
p_components <- ggplot(funnel_df, aes(x = x, y = y)) +
  geom_point(alpha = 0.1, size = 0.3, color = "gray") +
  labs(
    title = paste("GMM Components (", n_comp, "components)"),
    subtitle = "Component centers and 95% confidence ellipses",
    x = "x",
    y = "y"
  ) +
  theme_minimal()

# Add component centers and ellipses
for (k in 1:n_comp) {
  # Component center
  center <- compressed$means[, k]
  weight <- compressed$weights[k]
  sigma <- compressed$covariances[, , k]
  
  # Add center point
  p_components <- p_components +
    geom_point(
      data = data.frame(x = center[1], y = center[2]),
      aes(x = x, y = y),
      color = "red",
      size = 3,
      shape = 16
    )
  
  # Add ellipse (95% confidence)
  # Using ellipse package or manual calculation
  eigen_result <- eigen(sigma)
  angles <- atan2(eigen_result$vectors[2, 1], eigen_result$vectors[1, 1])
  chisq_val <- qchisq(0.95, df = 2)
  
  # Create ellipse points
  theta <- seq(0, 2 * pi, length.out = 100)
  ellipse_points <- cbind(
    cos(theta),
    sin(theta)
  ) %*% (eigen_result$vectors %*% diag(sqrt(eigen_result$values * chisq_val))) +
    matrix(center, nrow = 100, ncol = 2, byrow = TRUE)
  
  ellipse_df <- data.frame(
    x = ellipse_points[, 1],
    y = ellipse_points[, 2],
    component = k
  )
  
  p_components <- p_components +
    geom_path(
      data = ellipse_df,
      aes(x = x, y = y),
      color = "red",
      alpha = 0.5,
      linetype = "dashed"
    )
}

print(p_components)

# Component weights
weights_df <- data.frame(
  component = 1:n_comp,
  weight = compressed$weights
)

p_weights <- ggplot(weights_df, aes(x = component, y = weight)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Component Weights",
    x = "Component",
    y = "Weight"
  ) +
  theme_minimal()

print(p_weights)
```

# Summary

## Key Findings

1. **Compression Performance:**
   - Achieved `r round(compression_stats$compression_ratio[compression_stats$n_components == selected_n_comp], 1)`x compression ratio
   - Compression time: `r round(compression_stats$compression_time[compression_stats$n_components == selected_n_comp], 2)` seconds
   - Regeneration time: `r round(regeneration_time * 1000, 2)` milliseconds

2. **Quality of Approximation:**
   - The GMM with `r selected_n_comp` components captures the funnel shape reasonably well
   - Marginal distributions are well-preserved
   - Correlation structure is maintained

3. **Challenges:**
   - The funnel posterior is challenging due to its varying scales
   - More components may be needed for better approximation
   - The exponential relationship between x and y makes it difficult to capture with Gaussian components

## Recommendations

- Use more components (10-20) for complex distributions like the funnel
- Consider using "VVV" model (variable volume, shape, orientation) for maximum flexibility
- Evaluate BIC to balance model complexity and fit quality
- Visual inspection is important for assessing approximation quality

## Next Steps

- Try different covariance structures (EEE, VVE, etc.)
- Experiment with different numbers of components
- Compare with other compression methods
- Test on other challenging posterior distributions

```{r cleanup}
# Cleanup temporary files
unlink(csv_files)
unlink(file.path(temp_dir, "funnel_compressed_*.rds"))
cat("Cleaned up temporary files\n")
```

