---
title: "poco brms fit compression test"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author: 
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format: 
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Menu"
knitr:
  opts_chunk: 
    message: false
    warning: false
comments:
  hypothesis: 
    theme: clean
editor: source
abstract: |
  Compress a brms posterior end-to-end with the `poco` package, using a
  posteriorHCA brmsfit pulled from a `targets` pipeline.
---

# Library

```{r}
#| label: setup
.libPaths("/hpcfs/users/a1237163/R/x86_64-pc-linux-gnu-library/4.4")
dyn.load("/home/a1237163/miniconda3/lib/libglpk.so.40")
library(tidyverse)
library(tidySummarizedExperiment)
library(tidySingleCellExperiment)
library(tidybulk)
library(SingleCellExperiment)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(ggplot2)
library(scales)
library(tidytext)
library(glue)
library(targets)
library(here)
library(Seurat)
library(tidyseurat)
library(patchwork)
library(vissE)
library(tidyHeatmap)
library(circlize)
library(ggupset)
library(brms)
library(edgeR)
library(magrittr)
# set_cmdstan_path('/hpcfs/users/a1237163/R/.cmdstan/cmdstan-2.36.0')
```

# Pre-load and setting

```{r}
# Use the local poco source tree (loads R/, NAMESPACE, etc.) so we don't
# depend on an installed copy. Switch to `library(poco)` once installed.
pkgload::load_all("/home/a1237163/lab/chen/poco")

ct <- 
  'plasma' %>% 
  paste0('V1_',.)

target_pipeline_path = '/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/NEW_CELL_TYPE/'
target_obj_path = file.path(target_pipeline_path, ct, '_targets')

```


# Read brms fit file from pipeline
```{r}
set.seed(2603)

estimates_chunk_meta <- 
  tar_meta(
    starts_with('estimates_chunk_'),
    store = target_obj_path
  )


estimates_chunk <-
  tar_read_raw(
    estimates_chunk_meta$name %>% sample(1),
    store = target_obj_path
  )

fit_brms <- estimates_chunk$brms_fit[[1]]
  
```

```{r}
class(fit_brms)
print(object.size(fit_brms), units = "auto")
# summary(fit_brms)
```


## Inspect the Original Fit

```{r inspect_fit}
# Model summary
# summary(fit_brms)

# Check convergence
cat("\nConvergence diagnostics:\n")
cat("Max Rhat:", max(rhat(fit_brms), na.rm = TRUE), "\n")
cat("Min Bulk ESS:", min(neff_ratio(fit_brms) * 4000, na.rm = TRUE), "\n")

# Original posterior structure
original_draws <- as_draws_matrix(fit_brms)
cat("\nOriginal posterior dimensions:", dim(original_draws), "\n")
cat("Parameters:", colnames(original_draws)[1:10], "...\n")
```

# Posterior dependence structure

Before choosing a compression family we inspect the *posterior correlation*
structure. The goal is to identify several **correlated clusters** of
parameters (each deserves a non-diagonal mixture component) plus a
**remainder** block (unclustered / weakly correlated parameters) that can
be approximated with a cheaper diagonal / independent model.

## 1. Compute the correlation matrix

```{r posterior_corr}
posterior_cor <- posterior_correlation(
  original_draws,
  type   = "correlation",
  method = "pearson"
)

cat("Correlation matrix dim :", dim(posterior_cor), "\n")
cat("n_draws used           :", attr(posterior_cor, "n_draws"), "\n")
cat("Dropped (zero-variance):",
    length(attr(posterior_cor, "dropped_zero_var")), "\n")
```

## 2. Visual diagnostics for choosing a threshold

```{r posterior_corr_plots, fig.width=11, fig.height=8}
# A reasonable starting threshold; iterate after looking at the plots.
corr_threshold <- 0.25

corr_plots <- plot_posterior_correlation(
  posterior_cor,
  which              = c("hist", "max_abs"),
  threshold          = corr_threshold
)

# Distribution of |corr| over all parameter pairs + per-parameter
# "max |corr| with any other" curve, both with the threshold drawn.
(corr_plots$hist | corr_plots$max_abs)
```

## 3. Partition parameters into multiple clusters + a remainder block

`partition_parameters_clusters()` cuts the same hierarchical clustering
that the heatmap uses (`1 - |corr|` distance, average linkage) and keeps
only clusters with at least `min_size` members. **`min_size`** can be a
positive integer (minimum cluster count) or a **numeric proportion in
(0, 1)**, in which case the cutoff is `ceiling(min_size * n_parameters)`;
an informative line is printed when a proportion is used. Parameters not
in any kept cluster are collected in **`remainder`** (deprecated alias
`rest`).

```{r posterior_partition_clusters}
#| message: true
posterior_partition <- partition_parameters_clusters(
  posterior_cor,
  threshold = corr_threshold,
  # Integer: absolute minimum cluster size, or proportion in (0, 1):
  # min_size  = 0.02
  # e.g. 0.02 * 1156 params -> ceiling -> 24 (with full correlation matrix)
  min_size = 2L,
  # Optional: keep nuisance / dense correlation groups in the cheap remainder
  # (tidyselect OK), e.g. force_remainder = tidyselect::starts_with("cor_")
  # Or: force_remainder = c("param1", "param2")
  # Or specify a target number of clusters:
  #   k = 6, min_size = 10L
)

print(posterior_partition)

# Convenience handles for downstream blockwise compression.
posterior_blocks    <- posterior_partition$blocks     # named list of clusters
posterior_remainder <- posterior_partition$remainder  # unclustered tail
cat("\n# of correlated blocks :", length(posterior_blocks), "\n")
cat("Total in blocks         :", sum(lengths(posterior_blocks)), "\n")
cat("Remainder size          :", length(posterior_remainder), "\n")
```

## 4. Heatmap with row/column dendrograms and clusters outlined

The heatmap below reuses hierarchical clustering, draws **dendrograms on the
top and left**, and marks each kept cluster with a **coloured outline** (and
**on-plot labels when there is room**; with many clusters or dense grids,
labels and the cluster colour legend are omitted automatically—override with
`show_cluster_labels` / `show_cluster_legend`). Parameters in **`remainder`**
appear between marked regions (no outline).

```{r posterior_corr_heatmap, fig.width=11, fig.height=10}
heatmap_plot <- plot_posterior_correlation(
  posterior_cor,
  which              = "heatmap",
  max_params_heatmap = 300,
  dendrogram         = "both",
  partition          = posterior_partition
  # Uncomment to force on-plot cluster text even when the default hides it:
  # , show_cluster_labels = TRUE, show_cluster_legend = TRUE
)

heatmap_plot 

```

# Compression

`poco::compress_brmsfit()` returns a list with `$compressed` (the
`posterior_compressed` object) and `$structure` (the brmsfit shell with
draws stripped, ready for `reconstruct_brmsfit()`).

## A. Single-block compression (whole posterior at once)

```{r}
# Common mclust covariance structures:
#   "EII" - spherical, equal volume
#   "EEI" - diagonal, equal volume and shape
#   "VVI" - diagonal, variable volume and shape (compact for high-dim)
# Leave model_name = NULL to let poco/mclust pick by BIC.
fit_brms_compressed <- compress_brmsfit(
  brmsfit      = fit_brms,
  method       = "mclust",
  n_components = 3
  # model_name = "VVI",
)
```

```{r}
print(object.size(fit_brms_compressed), units = "auto")
fit_brms_compressed$compressed
print(object.size(fit_brms_compressed$compressed), units = "auto")
fit_brms_compressed$structure
print(object.size(fit_brms_compressed$structure), units = "auto")
```

## B. Hybrid blockwise compression (richer cov per cluster, diagonal for remainder)

Pass the `posterior_partition` from above as `partition`. Each cluster is
compressed independently (typically with a richer covariance family;
`cluster_model_name = NULL` lets mclust pick the best by BIC subject to
per-block identifiability), and the **remainder** block is compressed with
a cheap diagonal model (`remainder_model_name = "VVI"` by default).

The resulting `posterior_compressed_blockwise` object plugs into
`sample_posterior()` and `reconstruct_brmsfit()` exactly like a single-block
compression: per-block samples are recombined and re-aligned with the
original parameter order automatically.

This implies a **between-block independence approximation**: within-block
correlations are preserved, between-block correlations are dropped.

You can pass **`partition_parameters_clusters()`** output, a **plain list of
character vectors** (one block per list element; all other parameters go to
remainder), or **`partition_blocks()`** to mix literals with tidy-select
(e.g. `tidyselect::starts_with("b_")`, `tidyselect::contains("dataset_id")`).

Before fitting, **`compress_brmsfit()` / `compress_posterior()`** align the
partition to the actual posterior columns: names that appear only in the
partition are **dropped** with a **warning**; draw columns that appear in
neither a cluster nor `remainder` are **added to remainder** with a
**warning** (e.g. a smaller `variables` subset than the correlation matrix,
or new parameters in the fit).

**Parallel cluster fits (default):** with **`cluster_BPPARAM = NULL`**, poco
uses **BiocParallel** when it is installed and there are at least two cluster
blocks—worker count is **`min(n_clusters, parallel::detectCores() - 1)`**
(`SnowParam` on Windows, `MulticoreParam` elsewhere). If BiocParallel is not
installed, a **message** suggests `BiocManager::install("BiocParallel")` and
cluster blocks are compressed **sequentially** (no error). Set
**`cluster_BPPARAM = FALSE`** to force sequential fits even when BiocParallel
is available. Pass a **`BiocParallelParam`** (e.g. `BiocParallel::MulticoreParam(workers = 8L)`)
to override auto worker choice. The **remainder** block is always fitted in the
main process after clusters finish.

Set **`verbose = TRUE`** to print whether cluster blocks run in parallel, how
many **BiocParallel** workers are used, and (when supported) enable the
**progress bar** on the parallel job; per-cluster console lines are omitted
under parallel to avoid interleaved output, but the **remainder** block still
announces when `verbose` is `TRUE`. Use **`verbose = FALSE`** (default) for
silent runs.

```{r}
fit_brms_compressed_blockwise <- compress_brmsfit(
  brmsfit              = fit_brms,
  method               = "mclust",
  n_components         = 3,
  partition            = posterior_partition,

  # Manual blocks example:
  # partition            = list(c("param_a", "param_b"), c("param_c")),
  # partition            = partition_blocks(
  #   tidyselect::starts_with("b_"),
  #   tidyselect::contains("dataset_id")
  # ),
  verbose           = TRUE,   # parallel/sequential summary, workers, BiocParallel progress bar (+ backend chatter)
  
  # cluster_BPPARAM   = FALSE   # force sequential cluster blocks
  cluster_BPPARAM   = BiocParallel::MulticoreParam(workers = 8L)
  
  # cluster_model_name   = NULL,    # auto-pick richer family per cluster
  # remainder_model_name = "VVI"    # axis-aligned diagonal for remainder
)

fit_brms_compressed_blockwise$compressed
```

```{r}
print(object.size(fit_brms_compressed_blockwise$compressed), units = "auto")
print(object.size(fit_brms_compressed_blockwise), units = "auto")
```

Save compressed 
```{r}
compressed_file <- "brms_posterior_compressed_blockwise.rds"
saveRDS(fit_brms_compressed_blockwise$compressed, compressed_file, compress = "xz")


compressed_size <- file.size(compressed_file)
cat("\nCompressed size (with xz):", format(compressed_size, big.mark = ","), "bytes\n")
cat("That's", round(compressed_size / 1024, 2), "KB\n")
```

# Step 4: Regenerate Posterior Samples

Sample from the compressed posterior distribution.

```{r regenerate_samples}
# Generate same number of samples as original
n_draws_original <- nrow(original_draws)
cat("Generating", n_draws_original, "samples from compressed posterior...\n")

# poco::sample_posterior() accepts either an in-memory compressed object
# or a path to an `.rds` file containing one.
regenerated_samples <- sample_posterior(
  fit_brms_compressed$compressed,
  n_draws = n_draws_original
)

cat("Regenerated samples dimensions:", dim(regenerated_samples), "\n")
cat("Parameters:", colnames(regenerated_samples)[1:10], "...\n")

param_names <- fit_brms_compressed$compressed$param_names

# Basic quality check: compare means and SDs
original_means <- colMeans(original_draws[, param_names])
regenerated_means <- colMeans(regenerated_samples)

original_sds <- apply(original_draws[, param_names], 2, sd)
regenerated_sds <- apply(regenerated_samples, 2, sd)

cat("\nQuality check - first 5 parameters:\n")
comparison_df <- data.frame(
  Parameter = param_names[1:5],
  Original_Mean = round(original_means[1:5], 3),
  Regenerated_Mean = round(regenerated_means[1:5], 3),
  Original_SD = round(original_sds[1:5], 3),
  Regenerated_SD = round(regenerated_sds[1:5], 3)
)
print(comparison_df)
```

Now do the same from the **blockwise** compression (clusters use
non-diagonal mclust families, remainder uses diagonal). Parameter order
matches `param_names` automatically because `sample_posterior()` realigns
per-block samples to the original column order.

```{r regenerate_samples_blockwise}
regenerated_samples_blockwise <- sample_posterior(
  fit_brms_compressed_blockwise$compressed,
  n_draws = n_draws_original
)

cat("Blockwise samples dimensions :",
    dim(regenerated_samples_blockwise), "\n")
cat("Same parameter order         :",
    identical(colnames(regenerated_samples_blockwise), param_names), "\n")

regenerated_means_blockwise <- colMeans(
  regenerated_samples_blockwise[, param_names, drop = FALSE]
)
regenerated_sds_blockwise <- apply(
  regenerated_samples_blockwise[, param_names, drop = FALSE], 2, sd
)
```


# Visualization: Compare Original, Whole and Blockwise Compression

With ~`r length(param_names)` parameters we cannot eyeball every marginal,
so the plots below summarise the comparison along three complementary axes,
each one shown for **two regeneration methods** at once
(`whole` = single GMM over all parameters, `blockwise` = per-cluster GMM
+ diagonal remainder):

1. **Parameter-wide calibration** (one dot per parameter) of posterior means
   and SDs.
2. **Marginal density overlays** for a random subset of parameters.
3. **Joint 2D density** for random parameter pairs, to check whether the
   compressed mixture preserves dependencies and not just marginals.
4. **Per-parameter Q-Q** to assess tail fidelity.

```{r viz_setup}
# Color palette + small helper, kept local to this section.
src_colors <- c(
  original  = "steelblue",
  whole     = "darkgreen",
  blockwise = "darkorange"
)
src_levels <- names(src_colors)

# Pre-build the three aligned matrices we visualize repeatedly.
orig_mat        <- as.matrix(original_draws[, param_names, drop = FALSE])
regen_mat       <- as.matrix(regenerated_samples[, param_names, drop = FALSE])
regen_block_mat <- as.matrix(
  regenerated_samples_blockwise[, param_names, drop = FALSE]
)

# Drop posterior classes (e.g. draws_matrix) so tidyverse can combine columns.
orig_mat        <- unclass(orig_mat)
regen_mat       <- unclass(regen_mat)
regen_block_mat <- unclass(regen_block_mat)
storage.mode(orig_mat)        <- "double"
storage.mode(regen_mat)       <- "double"
storage.mode(regen_block_mat) <- "double"
```

## 1. Mean and SD calibration across all parameters

A perfect compression sits exactly on the red identity line; deviations reveal
which parameters are biased or over/under-dispersed by the GMM approximation.

```{r calibration_plots, fig.width=11, fig.height=5}
# One row per (parameter, method) so we can colour-code by method.
calibration_df <- dplyr::bind_rows(
  tibble::tibble(
    parameter     = param_names,
    method        = "whole",
    original_mean = original_means,
    regen_mean    = regenerated_means,
    original_sd   = original_sds,
    regen_sd      = regenerated_sds
  ),
  tibble::tibble(
    parameter     = param_names,
    method        = "blockwise",
    original_mean = original_means,
    regen_mean    = regenerated_means_blockwise,
    original_sd   = original_sds,
    regen_sd      = regenerated_sds_blockwise
  )
) |>
  dplyr::mutate(method = factor(method, levels = c("whole", "blockwise")))

reg_cols <- src_colors[c("whole", "blockwise")]

p_mean_calib <- ggplot(calibration_df,
                       aes(x = regen_mean, y = original_mean,
                           color = method)) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  scale_color_manual(values = reg_cols) +
  labs(
    title    = "Posterior means",
    subtitle = paste(length(param_names), "parameters"),
    x        = "Regenerated posterior mean",
    y        = "Original posterior mean"
  ) +
  theme_minimal()

# Robust axis limits so extreme outliers do not collapse all points.
mean_xy <- c(calibration_df$original_mean, calibration_df$regen_mean)
mean_lim <- stats::quantile(mean_xy, probs = c(0.01, 0.99), na.rm = TRUE)

p_mean_calib <- p_mean_calib +
  coord_equal(xlim = mean_lim, ylim = mean_lim)

p_sd_calib <- ggplot(calibration_df,
                     aes(x = regen_sd, y = original_sd,
                         color = method)) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  scale_color_manual(values = reg_cols) +
  labs(
    title    = "Posterior SDs",
    subtitle = paste(length(param_names), "parameters"),
    x        = "Regenerated posterior SD",
    y        = "Original posterior SD"
  ) +
  theme_minimal()

sd_xy <- c(calibration_df$original_sd, calibration_df$regen_sd)
sd_lim <- stats::quantile(sd_xy, probs = c(0.01, 0.99), na.rm = TRUE)
sd_lim[1] <- max(0, sd_lim[1])

p_sd_calib <- p_sd_calib +
  coord_equal(xlim = sd_lim, ylim = sd_lim)

(p_mean_calib + p_sd_calib) +
  patchwork::plot_layout(guides = "collect") &
  ggplot2::theme(legend.position = "bottom")
```

```{r calibration_summary}
calibration_summary <- calibration_df |>
  dplyr::group_by(method) |>
  dplyr::summarise(
    `RMSE(mean)`                 = sqrt(mean((regen_mean - original_mean)^2)),
    `MAE(mean)`                  = mean(abs(regen_mean - original_mean)),
    `median |relative SD error|` = median(abs(
      (regen_sd - original_sd) / pmax(original_sd, .Machine$double.eps)
    )),
    `max |relative SD error|`    = max(abs(
      (regen_sd - original_sd) / pmax(original_sd, .Machine$double.eps)
    )),
    .groups = "drop"
  )
print(calibration_summary)
```

## 2. Marginal density overlays for a random subset

```{r marginal_overlays, fig.width=11, fig.height=8}
set.seed(42)
n_show <- 9
sampled_params <- sample(param_names, min(n_show, length(param_names)))

long_subset <- function(mat, source) {
  tibble::as_tibble(mat[, sampled_params, drop = FALSE]) |>
    tidyr::pivot_longer(
      dplyr::everything(),
      names_to = "parameter", values_to = "value"
    ) |>
    dplyr::mutate(source = source)
}

marginal_df <- dplyr::bind_rows(
  long_subset(orig_mat,        "original"),
  long_subset(regen_mat,       "whole"),
  long_subset(regen_block_mat, "blockwise")
) |>
  dplyr::mutate(
    parameter = factor(parameter, levels = sampled_params),
    source    = factor(source, levels = src_levels)
  )

ggplot(marginal_df, aes(x = value, fill = source, color = source)) +
  geom_density(alpha = 0.30) +
  facet_wrap(~ parameter, scales = "free", ncol = 3) +
  scale_fill_manual(values = src_colors) +
  scale_color_manual(values = src_colors) +
  labs(
    title    = "Marginal posterior densities",
    subtitle = paste("Random subset of", length(sampled_params), "parameters"),
    x        = "Parameter value",
    y        = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 3. Joint 2D densities for random parameter pairs

This checks whether each compression preserves *dependencies* and not just
marginals. A diagonal-covariance mixture (e.g. `EEI`/`VVI`) cannot represent
within-component correlations, so off-diagonal "tilts" in original pairs may
be lost — by overlaying **original / whole / blockwise** we can see whether
the blockwise hybrid recovers tilts that the single-block (typically `VVI`)
compression flattened.

```{r joint_density, fig.width=11, fig.height=10}
set.seed(7)
n_pairs  <- 4
pair_idx <- replicate(n_pairs, sample(seq_along(param_names), 2),
                      simplify = FALSE)

pair_plots <- lapply(pair_idx, function(idx) {
  pp <- param_names[idx]
  df <- dplyr::bind_rows(
    tibble::tibble(x = orig_mat[, pp[1]],        y = orig_mat[, pp[2]],
                   source = "original"),
    tibble::tibble(x = regen_mat[, pp[1]],       y = regen_mat[, pp[2]],
                   source = "whole"),
    tibble::tibble(x = regen_block_mat[, pp[1]], y = regen_block_mat[, pp[2]],
                   source = "blockwise")
  ) |>
    dplyr::mutate(source = factor(source, levels = src_levels))

  ggplot(df, aes(x = x, y = y, color = source)) +
    geom_point(alpha = 0.18, size = 0.4) +
    stat_density_2d(alpha = 0.7, linewidth = 0.4) +
    scale_color_manual(values = src_colors) +
    labs(x = pp[1], y = pp[2]) +
    theme_minimal() +
    theme(axis.title = element_text(size = 8))
})

patchwork::wrap_plots(pair_plots, ncol = 2, guides = "collect") +
  patchwork::plot_annotation(
    title    = "Joint posteriors for random parameter pairs",
    subtitle = "Original vs whole-compression vs blockwise"
  ) &
  ggplot2::theme(legend.position = "bottom")
```

```{r correlation_calibration, fig.width=6, fig.height=6}
# Quantitative companion to the joint plots: how well are pairwise correlations
# preserved? We sample a few hundred random pairs and compare correlations.
set.seed(11)
n_corr_pairs <- min(500, choose(length(param_names), 2))
corr_pair_idx <- replicate(
  n_corr_pairs,
  sample(seq_along(param_names), 2),
  simplify = FALSE
)

corr_df <- purrr::map_dfr(corr_pair_idx, function(idx) {
  pp <- param_names[idx]
  tibble::tibble(
    pair            = paste(pp, collapse = " ~ "),
    cor_orig        = cor(orig_mat[, pp[1]],         orig_mat[, pp[2]]),
    cor_whole       = cor(regen_mat[, pp[1]],        regen_mat[, pp[2]]),
    cor_blockwise   = cor(regen_block_mat[, pp[1]],  regen_block_mat[, pp[2]])
  )
}) |>
  tidyr::pivot_longer(
    cols      = c("cor_whole", "cor_blockwise"),
    names_to  = "method",
    values_to = "cor_regen"
  ) |>
  dplyr::mutate(
    method = factor(
      sub("^cor_", "", method),
      levels = c("whole", "blockwise")
    )
  )

ggplot(corr_df, aes(x = cor_orig, y = cor_regen, color = method)) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.9) +
  scale_color_manual(values = src_colors[c("whole", "blockwise")]) +
  facet_wrap(~ method) +
  coord_equal(xlim = c(-1, 1), ylim = c(-1, 1)) +
  labs(
    title    = "Pairwise correlations: original vs regenerated",
    subtitle = paste(
      length(unique(corr_df$pair)),
      "random parameter pairs (per method)"
    ),
    x        = "cor(original)",
    y        = "cor(regenerated)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## 4. Per-parameter Q-Q plots

Tail behaviour matters for predictive intervals. The Q-Q points should hug
the red identity line if the GMM captures the full distribution (not just
mean/SD).

```{r qq_plots, fig.width=11, fig.height=7}
qq_params <- sampled_params[seq_len(min(6, length(sampled_params)))]
qq_probs  <- ppoints(200)

qq_df <- purrr::map_dfr(qq_params, function(p) {
  q_orig <- stats::quantile(orig_mat[, p],        probs = qq_probs, names = FALSE)
  dplyr::bind_rows(
    tibble::tibble(
      parameter = p, method = "whole",
      q_orig    = q_orig,
      q_regen   = stats::quantile(regen_mat[, p],
                                  probs = qq_probs, names = FALSE)
    ),
    tibble::tibble(
      parameter = p, method = "blockwise",
      q_orig    = q_orig,
      q_regen   = stats::quantile(regen_block_mat[, p],
                                  probs = qq_probs, names = FALSE)
    )
  )
}) |>
  dplyr::mutate(
    parameter = factor(parameter, levels = qq_params),
    method    = factor(method,    levels = c("whole", "blockwise"))
  )

ggplot(qq_df, aes(x = q_orig, y = q_regen, color = method)) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.6, size = 0.7) +
  scale_color_manual(values = src_colors[c("whole", "blockwise")]) +
  facet_wrap(~ parameter, scales = "free", ncol = 3) +
  labs(
    title    = "Q-Q plots: original vs regenerated quantiles",
    subtitle = "Points on the red line indicate good distributional fidelity",
    x        = "Original quantiles",
    y        = "Regenerated quantiles",
    color    = "method"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


# Reconstruct brms fit objects

`reconstruct_brmsfit()` regenerates draws from a compressed posterior and
re-injects them into the saved brms shell so downstream brms methods
(`predict()`, `posterior_predict()`, `pp_check()`, `conditional_effects()`,
...) keep working. The exact same call works for both single-block and
blockwise compressions: per-block samples are realigned to the original
parameter order before the per-chain refill.

Sizes printed below should be much smaller than the original `fit_brms`
because both reconstructed objects carry only the regenerated chains
(no warm-up, no sampler diagnostics).

```{r reconstruct_brmsfit_both}
fit_brms_recon_whole <- reconstruct_brmsfit(
  fit_brms_compressed,
  n_draws = n_draws_original
)

fit_brms_recon_blockwise <- reconstruct_brmsfit(
  fit_brms_compressed_blockwise,
  n_draws = n_draws_original
)

# Sanity check: both reconstructions are valid brmsfit objects.
stopifnot(inherits(fit_brms_recon_whole,     "brmsfit"))
stopifnot(inherits(fit_brms_recon_blockwise, "brmsfit"))

cat("compression_method (whole)     :",
    attr(fit_brms_recon_whole, "compression_method"), "\n")
cat("compression_method (blockwise) :",
    attr(fit_brms_recon_blockwise, "compression_method"), "\n")
cat("compression_base_method        :",
    attr(fit_brms_recon_blockwise, "compression_base_method"), "\n")
cat("\nblock sizes (blockwise reconstruction):\n")
print(attr(fit_brms_recon_blockwise, "compression_blocks"))
```

## Object size comparison

```{r reconstruct_size_compare}
size_tbl <- tibble::tibble(
  object = c(
    "fit_brms (original)",
    "fit_brms_compressed$compressed",
    "fit_brms_compressed_blockwise$compressed",
    "fit_brms_recon_whole",
    "fit_brms_recon_blockwise"
  ),
  bytes = c(
    as.numeric(object.size(fit_brms)),
    as.numeric(object.size(fit_brms_compressed$compressed)),
    as.numeric(object.size(fit_brms_compressed_blockwise$compressed)),
    as.numeric(object.size(fit_brms_recon_whole)),
    as.numeric(object.size(fit_brms_recon_blockwise))
  )
) |>
  dplyr::mutate(
    pretty = vapply(
      bytes,
      function(b) format(structure(b, class = "object_size"), units = "auto"),
      character(1L)
    )
  )

print(size_tbl)
```

## Quick verification: draws survive the round-trip

Pull the user-facing draws back via `posterior::as_draws_matrix()` from each
reconstructed brmsfit and confirm dimensions and parameter alignment match
the original.

```{r reconstruct_round_trip}
draws_recon_whole     <- as_draws_matrix(fit_brms_recon_whole)
draws_recon_blockwise <- as_draws_matrix(fit_brms_recon_blockwise)

cat("Original draws            :", dim(original_draws),          "\n")
cat("Reconstructed (whole)     :", dim(draws_recon_whole),       "\n")
cat("Reconstructed (blockwise) :", dim(draws_recon_blockwise),   "\n")
cat(
  "Same param order (whole)     : ",
  identical(colnames(draws_recon_whole),     param_names), "\n",
  "Same param order (blockwise) : ",
  identical(colnames(draws_recon_blockwise), param_names), "\n",
  sep = ""
)
```

