---
title: "Compressing sccomp posteriors"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Compressing sccomp posteriors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

This vignette walks through compressing an `sccomp` fit, saving the
compressed posterior, and reconstructing an analysis-ready object.
It requires a working `sccomp` + `cmdstanr` setup.

```{r libs}
library(poco)
```

## Fit a sccomp model

`compress_sccomp()` expects an `sccomp` object that carries the
underlying `cmdstanr` fit in `attr(x, "fit")`. Here we use the
`counts_obj` example dataset shipped with `sccomp`.

```{r fit_sccomp}
library(sccomp)
library(dplyr)

data("counts_obj", package = "sccomp")
draws_dir <- file.path(tempdir(), "sccomp_draws_files")
dir.create(draws_dir, recursive = TRUE, showWarnings = FALSE)

fit <-
  counts_obj |>
  sccomp_estimate(
    formula_composition = ~ type,
    sample              = "sample",
    cell_group          = "cell_group",
    abundance           = "count",
    inference_method    = "hmc",
    portable            = FALSE,
    output_directory    = draws_dir,
    cores               = 1,
    verbose             = FALSE
  )
```

## Compress and save

`compress_sccomp()` returns a list with two parts:

- `compressed`: compact posterior representation
- `structure`: the original `sccomp` object, used during reconstruction

```{r compress}
result <- compress_sccomp(
  fit,
  method       = "mvdens_gmm",
  n_components = 8
)
```

Here we keep `portable = FALSE` and set an explicit `output_directory`,
so cmdstanr CSV files remain available and `compress_sccomp()` can read
all posterior parameters.

## Reconstruct and continue

```{r reconstruct}
fit_recon <- reconstruct_sccomp(result)

fit_recon
```

## Compare pairwise posterior geometry

Use `bayesplot::mcmc_pairs()` to compare pairwise structure before and
after reconstruction. The original draws come from the `cmdstanr` fit
attached to the `sccomp` object; the reconstructed draws are the
regenerated samples stored on the reconstructed object.

```{r pairs}
library(bayesplot)

fit_cmdstan <- attr(fit, "fit")

regen_samples <- attr(fit_recon, "fit_compressed")$regenerated_samples
pair_pars <- head(colnames(regen_samples), 4)

draws_orig  <- fit_cmdstan$draws(variables = pair_pars, format = "draws_array")
draws_recon <- posterior::as_draws_array(
  regen_samples[, pair_pars, drop = FALSE]
)

bayesplot::mcmc_pairs(draws_orig,  pars = pair_pars)
bayesplot::mcmc_pairs(draws_recon, pars = pair_pars)
```

MCMC diagnostics (`Rhat`, ESS) are *not* preserved by the regenerated
draws.

## Other methods

`compress_sccomp()` shares the same `method` argument as
`compress_posterior()`:

```{r methods}
compression_methods()

result_kde <- compress_sccomp(
  fit,
  method    = "mvdens_kde"
)
```

## Session info

```{r si, eval = TRUE}
sessionInfo()
```
