---
title: "Compressing brms posteriors"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Compressing brms posteriors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  eval = FALSE
)
```

This vignette walks through the full workflow for compressing a
`brms::brmsfit`, then reconstructing it for use with the standard
`brms` post-processing functions (`predict()`, `posterior_predict()`,
`pp_check()`, ...). Chunks are not evaluated by default because they
require `brms` + `cmdstanr` + a working CmdStan install.

```{r libs, eval = TRUE}
library(poco)
```

## Fit a brms model

The model **must** use the cmdstanr backend so that
`compress_brmsfit()` can reach the CSV draws.

```{r fit_brms}
library(brms)
library(dplyr)

set.seed(1)
n_groups <- 20; n_per <- 15
dat <- expand.grid(group = 1:n_groups, rep = 1:n_per)
dat$x  <- rnorm(nrow(dat))
re     <- rnorm(n_groups, 0, 0.5)
dat$y  <- 2 + 1.5 * dat$x + re[dat$group] + rnorm(nrow(dat), 0, 0.8)

fit <- brm(
  y ~ x + (1 | group),
  data    = dat,
  backend = "cmdstanr",
  chains  = 4,
  iter    = 2000,
  warmup  = 1000,
  refresh = 0
)
```

## Compress and save

A single call returns both the compressed posterior and the brms fit
structure. Save them separately so the structure can be reused with any
future updates to the compressed posterior.

```{r compress}
result <- compress_brmsfit(
  fit,
  method       = "mclust",
  n_components = 5,
  remove_csvs  = FALSE
)

saveRDS(result$compressed, "model_compressed.rds", compress = "xz")
saveRDS(result$structure,  "model_structure.rds")
```

`result$compressed` carries the GMM parameters; `result$structure` is
the original `brmsfit`, which we keep so that downstream methods can be
re-attached.

## Reconstruct and use the model

```{r reconstruct}
result_loaded <- list(
  compressed = readRDS("model_compressed.rds"),
  structure  = readRDS("model_structure.rds")
)

fit_recon <- reconstruct_brmsfit(result_loaded)

# Now use brms machinery as normal:
new_data <- data.frame(x = seq(-2, 2, length.out = 20),
                       group = rep(1:5, each = 4))

predict(fit_recon, newdata = new_data)
posterior_predict(fit_recon, newdata = new_data)
conditional_effects(fit_recon)
```

MCMC diagnostics (`Rhat`, ESS) are *not* preserved by the regenerated
draws.

## Other methods

`compress_brmsfit()` shares the same `method` argument as
`compress_posterior()`. To use a different backend, e.g. mvdens GMM:

```{r mvdens}
result <- compress_brmsfit(
  fit,
  method       = "mvdens_gmm",
  n_components = 5
)
```

(Requires the suggested `mvdens` package -
`remotes::install_github("NKI-CCB/mvdens")`.)

## Session info

```{r si, eval = TRUE}
sessionInfo()
```
