---
title: "Compressing a cmdstanr funnel posterior"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Compressing a cmdstanr funnel posterior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

This vignette compresses Neal's funnel posterior (a classic stress test
because of the strong dependency between `y` and `x`) with `compress_fit()`,
then regenerates samples and inspects the approximation.

It requires a working CmdStan install.

```{r libs}
library(poco)
```

## Fit the funnel model

```{r fit_funnel}
library(cmdstanr)

stan_code <- "
data { int<lower=1> D; }
parameters {
  real y;
  vector[D] x;
}
model {
  y ~ normal(0, 3);
  x ~ normal(0, exp(y / 2));
}
"
mod <- cmdstan_model(write_stan_file(stan_code))

D <- 6L
fit <- mod$sample(
  data           = list(D = D),
  chains         = 4,
  parallel_chains = 4,
  iter_warmup    = 1000,
  iter_sampling  = 1000,
  refresh        = 0
)
```

## Compress

```{r compress}
vars <- c("y", paste0("x[", 1:D, "]"))
methods <- c("mclust", "mvdens_gmm", "mvdens_kde")

comps <- lapply(methods, function(m) {
  tryCatch(
    compress_fit(
      fit,
      variables    = vars,
      method       = m,
      n_components = 5
    ),
    error = function(e) e
  )
})
names(comps) <- methods

compress_status <- data.frame(
  method = methods,
  ok = vapply(comps, function(x) !inherits(x, "error"), logical(1)),
  message = vapply(
    comps,
    function(x) if (inherits(x, "error")) conditionMessage(x) else "ok",
    character(1)
  ),
  row.names = NULL
)
compress_status
```

You can also pass the CSV files directly (example shown for `mclust`):

```{r compress_csv}
comp_csv <- compress_fit(
  fit$output_files(),
  variables    = vars,
  method       = "mclust",
  n_components = 5
)
comp_csv
```

## Regenerate samples and compare

```{r regenerate}
draws_orig <- fit$draws(variables = vars, format = "matrix")
draws_new_list <- lapply(comps, function(comp) {
  if (inherits(comp, "error")) return(comp)
  tryCatch(
    sample_posterior(comp, n_draws = nrow(draws_orig)),
    error = function(e) e
  )
})

regen_status <- data.frame(
  method = names(draws_new_list),
  ok = vapply(draws_new_list, function(x) !inherits(x, "error"), logical(1)),
  message = vapply(
    draws_new_list,
    function(x) if (inherits(x, "error")) conditionMessage(x) else "ok",
    character(1)
  )
)
regen_status

ok_draws <- draws_new_list[vapply(draws_new_list, function(x) !inherits(x, "error"), logical(1))]

mean_summary <- do.call(
  rbind,
  c(list(original = colMeans(draws_orig)), lapply(ok_draws, colMeans))
)
round(mean_summary, 3)
```

```{r plots, message = FALSE, warning = FALSE}
library(ggplot2)

stacked <- do.call(
  rbind,
  c(
    list(data.frame(draws_orig[, c("y", "x[1]")], source = "original",
                    check.names = FALSE)),
    lapply(names(ok_draws), function(m) {
      data.frame(
        ok_draws[[m]][, c("y", "x[1]")],
        source = m,
        check.names = FALSE
      )
    })
  )
)

p1 <- ggplot(stacked, aes(y, fill = source)) +
  geom_density(alpha = 0.35) + theme_minimal()
p2 <- ggplot(stacked, aes(y, `x[1]`, color = source)) +
  geom_point(alpha = 0.2, size = 0.4) + theme_minimal()
patchwork::wrap_plots(p1, p2, ncol = 2)
```

## Session info

```{r si, eval = TRUE}
sessionInfo()
```
