---
title: "Compressing sccomp posteriors"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Compressing sccomp posteriors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  eval = FALSE
)
```

This vignette walks through compressing an `sccomp` fit, saving the
compressed posterior, and reconstructing an analysis-ready object.
Chunks are not evaluated by default because they require a working
`sccomp` + `cmdstanr` setup.

```{r libs, eval = TRUE}
library(poco)
```

## Fit a sccomp model

`compress_sccomp()` expects an `sccomp` object that carries the
underlying `cmdstanr` fit in `attr(x, "fit")`.

```{r fit_sccomp}
library(sccomp)
library(dplyr)

# Replace this with your own data and formula.
fit <- sccomp_glm(
  data                = your_data,
  formula_composition = ~ condition,
  .sample             = sample,
  .cell_group         = cell_group,
  .count              = count
)
```

## Compress and save

`compress_sccomp()` returns a list with two parts:

- `compressed`: compact posterior representation
- `structure`: the original `sccomp` object, used during reconstruction

```{r compress}
result <- compress_sccomp(
  fit,
  method       = "mclust",
  n_components = 5,
  remove_csvs  = FALSE
)

saveRDS(result$compressed, "sccomp_compressed.rds", compress = "xz")
saveRDS(result$structure,  "sccomp_structure.rds")
```

## Reconstruct and continue

```{r reconstruct}
result_loaded <- list(
  compressed = readRDS("sccomp_compressed.rds"),
  structure  = readRDS("sccomp_structure.rds")
)

fit_recon <- reconstruct_sccomp(result_loaded)

fit_recon
```

## Compare pairwise posterior geometry

Use `bayesplot::mcmc_pairs()` to compare pairwise structure before and
after reconstruction.

```{r pairs}
library(bayesplot)

fit_cmdstan <- attr(fit, "fit")
fit_recon_cmdstan <- attr(fit_recon, "fit")

all_pars <- fit_cmdstan$variables()
pair_pars <- head(all_pars[!grepl("__", all_pars)], 4)
if (length(pair_pars) < 2) {
  pair_pars <- head(all_pars, min(4, length(all_pars)))
}

draws_orig <- fit_cmdstan$draws(variables = pair_pars, format = "draws_array")
draws_recon <- fit_recon_cmdstan$draws(variables = pair_pars, format = "draws_array")

bayesplot::mcmc_pairs(draws_orig, pars = pair_pars)
bayesplot::mcmc_pairs(draws_recon, pars = pair_pars)
```

MCMC diagnostics (`Rhat`, ESS) are *not* preserved by the regenerated
draws.

## Other methods

`compress_sccomp()` shares the same `method` argument as
`compress_posterior()`:

```{r methods}
compression_methods()

result_kde <- compress_sccomp(
  fit,
  method = "mvdens_kde"
)
```

## Session info

```{r si, eval = TRUE}
sessionInfo()
```
