---
title: "Evaluating compression fidelity"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Evaluating compression fidelity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4,
  message    = FALSE,
  warning    = FALSE
)
```

Posterior compression is lossy by construction - just like JPEG for
images. `evaluate_compression()` reports a **reproduction percentage**
that quantifies how much information was lost, and it does so without
being *circular*: the score does not reuse the same density model
(`mclust`, `mvdens_gmm`, `mvdens_kde`) that performed the compression.

This vignette walks through the metric on three sanity checks where the
visual answer is obvious, then a sweep that pins down the metric's
behaviour as we increase the difficulty of the problem.

## Two non-circular yardsticks

`evaluate_compression()` combines two distribution-free, sample-based,
two-sample diagnostics:

- **Energy distance** (Szekely & Rizzo, 2013): a multivariate distance
  between two empirical samples that is zero iff the two distributions
  match. It is anchored against a *bootstrap noise envelope* obtained
  by resampling the reference draws, so a reproduction of 100% means
  the regenerated sample is statistically indistinguishable from a
  fresh draw of the reference.
- **Classifier two-sample test (C2ST)** (Lopez-Paz & Oquab, 2017): a
  classifier (random forest via `ranger`, falling back to k-nearest
  neighbours) is trained under cross-validation to discriminate
  reference from reconstructed draws. AUC = 0.5 means indistinguishable;
  the score is `100 * (1 - 2 * |AUC - 0.5|)`.

Both metrics work purely from samples returned by `sample_posterior()`;
they never inspect the GMM weights or KDE bandwidths.

## Setup

```{r libs}
library(poco)
library(ggplot2)
library(patchwork)
library(mvtnorm)

set.seed(1)
```

```{r helpers, include = FALSE}
plot_overlay <- function(reference, recon, title) {
  ref_df   <- data.frame(reference, source = "reference")
  recon_df <- data.frame(recon,     source = "reconstructed")
  df <- rbind(ref_df, recon_df)
  cols <- colnames(reference)
  ggplot(df, aes(x = .data[[cols[1]]], y = .data[[cols[2]]],
                 colour = source)) +
    geom_point(alpha = 0.3, size = 0.6) +
    scale_colour_manual(values = c(reference = "gray40",
                                   reconstructed = "firebrick")) +
    labs(title = title, x = cols[1], y = cols[2]) +
    theme_bw() +
    theme(legend.position = "bottom")
}
```

# Sanity check 1: a faithful compression

A bivariate Gaussian with strong correlation, compressed with a single
**full-covariance** Gaussian (the textbook right answer).

```{r demo1-data}
Sigma <- matrix(c(1.0, 0.7, 0.7, 1.0), 2, 2)
draws <- rmvnorm(2000, mean = c(2, -1), sigma = Sigma)
colnames(draws) <- c("alpha", "beta")
```

```{r demo1-fit}
comp_full <- compress_posterior(
  draws, method = "mclust",
  n_components = 1L, model_name = "VVV"
)
recon_full <- sample_posterior(comp_full, n_draws = 2000)

fid_full <- evaluate_compression(
  comp_full, draws,
  classifier = "knn", seed = 1L
)
fid_full
```

```{r demo1-plot, fig.width = 6, fig.height = 5, echo = FALSE}
plot_overlay(
  draws, recon_full,
  sprintf("Faithful (full covariance) - %.1f%% reproduction",
          fid_full$reproduction_pct)
)
```

The reconstructed cloud is visually indistinguishable from the
reference. Both metrics agree it is at the noise floor.

# Sanity check 2: the killer test - lost correlation

Now compress the **same** correlated posterior with a *diagonal-only*
mclust model. By construction, every component has zero off-diagonal
covariance, so any joint correlation between `alpha` and `beta` simply
cannot be represented.

```{r demo2-fit}
comp_diag <- compress_posterior(
  draws, method = "mclust",
  n_components = 1L,
  model_name = c("EII", "VII", "EEI", "VEI", "EVI", "VVI")
)
recon_diag <- sample_posterior(comp_diag, n_draws = 2000)

fid_diag <- evaluate_compression(
  comp_diag, draws,
  classifier = "knn", seed = 1L
)
fid_diag
```

```{r demo2-plot, fig.width = 11, fig.height = 5, echo = FALSE}
(plot_overlay(draws, recon_full,
              sprintf("Full covariance: %.1f%%",
                      fid_full$reproduction_pct)) +
 plot_overlay(draws, recon_diag,
              sprintf("Diagonal only: %.1f%%",
                      fid_diag$reproduction_pct)))
```

Visually, the orange cloud on the right is *circular* where the gray
reference is *elongated*: the reconstruction has lost the correlation.
The empirical Pearson correlations confirm what the eye sees:

```{r demo2-corr}
data.frame(
  source           = c("reference", "diagonal recon", "full recon"),
  cor_alpha_beta   = c(cor(draws)[1, 2],
                       cor(recon_diag)[1, 2],
                       cor(recon_full)[1, 2])
) |>
  transform(cor_alpha_beta = round(cor_alpha_beta, 3))
```

The reproduction score drops from `r round(fid_full$reproduction_pct)`%
to `r round(fid_diag$reproduction_pct)`% - and crucially, **both**
metrics independently catch it (energy
`r round(fid_diag$metrics$energy$reproduction_pct)`%, C2ST
`r round(fid_diag$metrics$c2st$reproduction_pct)`%):

```{r demo2-bothmetrics, echo = FALSE}
data.frame(
  metric                 = c("energy", "c2st", "headline"),
  full_covariance        = round(c(
    fid_full$metrics$energy$reproduction_pct,
    fid_full$metrics$c2st$reproduction_pct,
    fid_full$reproduction_pct
  ), 1),
  diagonal_only          = round(c(
    fid_diag$metrics$energy$reproduction_pct,
    fid_diag$metrics$c2st$reproduction_pct,
    fid_diag$reproduction_pct
  ), 1)
)
```

# Sanity check 3: a collapsed mode

A bimodal posterior, compressed with two or one components.

```{r demo3-data}
m1 <- rmvnorm(1500, mean = c(-2, -1),
              sigma = matrix(c(0.4, 0.0, 0.0, 0.3), 2, 2))
m2 <- rmvnorm(1500, mean = c( 2,  1),
              sigma = matrix(c(0.3, 0.0, 0.0, 0.5), 2, 2))
bi  <- rbind(m1, m2)
colnames(bi) <- c("alpha", "beta")
```

```{r demo3-fits}
comp_two <- compress_posterior(bi, method = "mclust", n_components = 2L)
comp_one <- compress_posterior(bi, method = "mclust", n_components = 1L)

recon_two <- sample_posterior(comp_two, n_draws = 3000)
recon_one <- sample_posterior(comp_one, n_draws = 3000)

fid_two <- evaluate_compression(comp_two, bi, classifier = "knn", seed = 1L)
fid_one <- evaluate_compression(comp_one, bi, classifier = "knn", seed = 1L)
```

```{r demo3-plot, fig.width = 11, fig.height = 5, echo = FALSE}
(plot_overlay(bi, recon_two,
              sprintf("2 components: %.1f%%",
                      fid_two$reproduction_pct)) +
 plot_overlay(bi, recon_one,
              sprintf("1 component (mode collapse): %.1f%%",
                      fid_one$reproduction_pct)))
```

The single-component reconstruction visibly *bridges* the two real
modes - and the score drops from
`r round(fid_two$reproduction_pct)`% to
`r round(fid_one$reproduction_pct)`%. Once again, both metrics agree
something is wrong.

# Sanity sweep: reproduction vs the difficulty of the problem

If the metric is doing its job, increasing the *true* correlation
should make a diagonal-only compression progressively worse. We sweep
the correlation from 0 to 0.95 and watch the reproduction score.

```{r sweep, fig.width = 8, fig.height = 4}
sweep_one_rho <- function(rho, n = 1500, seed = 7L) {
  set.seed(seed)
  S <- matrix(c(1, rho, rho, 1), 2, 2)
  d <- rmvnorm(n, sigma = S)
  colnames(d) <- c("alpha", "beta")
  comp <- compress_posterior(
    d, method = "mclust", n_components = 1L,
    model_name = c("EII", "VII", "EEI", "VEI", "EVI", "VVI")
  )
  fid <- evaluate_compression(
    comp, d,
    classifier  = "knn",
    seed        = 1L,
    n_self_reps = 10L,
    max_n       = 800L
  )
  data.frame(
    rho      = rho,
    energy   = fid$metrics$energy$reproduction_pct,
    c2st     = fid$metrics$c2st$reproduction_pct,
    headline = fid$reproduction_pct
  )
}

rhos <- seq(0, 0.95, by = 0.05)
sweep_df <- do.call(rbind, lapply(rhos, sweep_one_rho))

sweep_long <- rbind(
  data.frame(rho = sweep_df$rho, metric = "energy",
             pct = sweep_df$energy),
  data.frame(rho = sweep_df$rho, metric = "c2st",
             pct = sweep_df$c2st),
  data.frame(rho = sweep_df$rho, metric = "headline (mean)",
             pct = sweep_df$headline)
)

ggplot(sweep_long, aes(x = rho, y = pct, colour = metric)) +
  geom_hline(yintercept = 100, linetype = 3, colour = "gray70") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  scale_colour_manual(values = c(
    "energy"           = "steelblue",
    "c2st"             = "firebrick",
    "headline (mean)"  = "gray30"
  )) +
  scale_y_continuous(limits = c(0, 105),
                     breaks = seq(0, 100, by = 20)) +
  labs(
    title    = "Diagonal-only compression: fidelity drops as correlation grows",
    subtitle = "Both metrics independently track the lost joint structure",
    x        = expression("True correlation " * rho),
    y        = "reproduction (%)",
    colour   = "metric"
  ) +
  theme_bw()
```

Both energy and C2ST decrease monotonically with the strength of the
joint structure that the diagonal model cannot capture. At `rho = 0`
the diagonal model *is* the truth and reproduction sits at 100%; by
`rho = 0.95` the metric correctly reports a substantial loss.

# Strict out-of-sample evaluation

If you want to rule out any residual concern that the GMM has seen the
reference draws (it has, because we fit on the same `draws`), split
*before* fitting:

```{r holdout}
n   <- nrow(bi)
idx <- sample.int(n, size = floor(0.8 * n))
comp_train <- compress_posterior(
  bi[idx, ], method = "mclust", n_components = 2L
)

fid_holdout <- evaluate_compression(
  comp_train,
  reference_draws = bi[-idx, ],
  classifier      = "knn",
  seed            = 1L
)
fid_holdout
```

The interpretation is identical, but the reference is data the GMM
never saw. Useful as a final sign-off that the score is not
optimistically biased.

# Wrap-up

`evaluate_compression()` gives you:

- a **single, interpretable number** (`reproduction_pct`) suitable for
  reporting alongside compressed posteriors;
- two **independent** corroborating metrics (energy distance and
  C2ST), so a single high score is unlikely to be a fluke;
- a workflow that is **backend-agnostic**: the same function works
  identically for `mclust`, `mvdens_gmm`, and `mvdens_kde` outputs.

The sanity checks above are also a useful template for authors of new
compression backends to verify that a method has not silently lost
correlation, multimodality, or any other joint structure.

## Session info

```{r si}
sessionInfo()
```
