---
title: "Introduction to poco"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to poco}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6,
  fig.height = 4
)
```

`poco` (POsterior COmpression, "little" in Italian) provides a small, unified API for turning a posterior
draws matrix into a compact density approximation, then drawing fresh
samples or evaluating the density on demand.

This vignette shows the matrix-in / samples-out workflow on a synthetic
two-component example. See `vignette("compress-brms")` and
`vignette("compress-funnel")` for end-to-end workflows on real `brms` and
`cmdstanr` fits.

```{r libs}
library(poco)
library(mvtnorm)
```

## A toy posterior

We build a synthetic two-blob "posterior" so we can see compression
preserve multimodality.

```{r toy_data}
set.seed(1)
n_per <- 2000
draws <- rbind(
  mvtnorm::rmvnorm(n_per, mean = c(-2, -1),
                   sigma = matrix(c(0.5, 0.1, 0.1, 0.3), 2)),
  mvtnorm::rmvnorm(n_per, mean = c( 2,  1),
                   sigma = matrix(c(0.4, -0.2, -0.2, 0.6), 2))
)
colnames(draws) <- c("alpha", "beta")
head(draws)
```

## Compress

Pick a `method` (the rest of the API does not change):

```{r compress}
comp <- compress_posterior(
  draws,
  method = "mclust",
  n_components = 2
)
comp
```

`compression_methods()` lists the currently supported choices.

```{r methods}
compression_methods()
```

## Regenerate samples

```{r sample}
new_draws <- sample_posterior(comp, n_draws = 5000)
dim(new_draws)
head(new_draws)
```

A quick sanity check on means and covariances:

```{r quality}
round(colMeans(draws),     3); round(colMeans(new_draws),     3)
round(cov(draws),          3); round(cov(new_draws),          3)
```

## Evaluate the density

`density_posterior()` accepts a matrix of points (one per row).

```{r density}
test_pts <- rbind(
  c(-2, -1),
  c( 0,  0),
  c( 2,  1)
)
density_posterior(comp, test_pts)
density_posterior(comp, test_pts, log = TRUE)
```

## Save / load

The compressed object is a plain list, so `saveRDS()` (with xz
compression) works out of the box.

```{r save_load}
tf <- tempfile(fileext = ".rds")
saveRDS(comp, tf, compress = "xz")
file.size(tf)

comp_loaded <- readRDS(tf)
identical(class(comp_loaded), class(comp))

s <- sample_posterior(tf, n_draws = 100)  # path also works
dim(s)

unlink(tf)
```

## Where to next

- `vignette("compress-brms", package = "poco")` for a
  full `brms::brmsfit` round-trip including reconstruction.
- `vignette("compress-funnel", package = "poco")` for a
  cmdstanr workflow on Neal's funnel.

## Session info

```{r si}
sessionInfo()
```
