---
title: "Comparing compression methods"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Comparing compression methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

`poco` exposes a single `method` argument to switch between density
approximations. This vignette compares the available backends on a
small synthetic posterior so you can pick the right trade-off for your
problem.

```{r libs}
library(poco)
library(mvtnorm)
```

## Available methods

```{r methods}
compression_methods()
```

| `method`        | Backend                    | Notes |
| --------------- | -------------------------- | ----- |
| `"mclust"`      | `mclust::Mclust()`         | parametric GMM, BIC-based selection, fast & compact |
| `"mvdens_gmm"`  | `mvdens::fit.gmm()`        | EM-based GMM, requires the suggested `mvdens` package |
| `"mvdens_kde"`  | `mvdens::fit.kde()`        | kernel density estimate, larger but non-parametric |

## A toy posterior with correlated parameters

```{r toy}
set.seed(1)
n_per <- 2000
draws <- rbind(
  mvtnorm::rmvnorm(
    n_per,
    mean = c(5.5, 2.0, 2.0),
    sigma = matrix(c(
      1.0, -0.14, 0.00,
      -0.14, 0.03, 0.00,
      0.00, 0.00, 0.05
    ), 3, 3, byrow = TRUE)
  ),
  mvtnorm::rmvnorm(
    n_per,
    mean = c(7.0, 1.8, 1.9),
    sigma = matrix(c(
      1.2, -0.18, 0.00,
      -0.18, 0.04, 0.00,
      0.00, 0.00, 0.06
    ), 3, 3, byrow = TRUE)
  )
)
colnames(draws) <- c("alpha", "beta", "sigma")
```

## Compress with each method

`mvdens_*` methods require the suggested `mvdens` package; we skip them
gracefully when it is not installed.

```{r compress}
methods_to_try <- c("mclust")
if (requireNamespace("mvdens", quietly = TRUE)) {
  methods_to_try <- c(methods_to_try, "mvdens_gmm", "mvdens_kde")
}

fits <- lapply(methods_to_try, function(m) {
  compress_posterior(draws, method = m, n_components = 2)
})
names(fits) <- methods_to_try
```

## Storage footprint

A practical proxy for "how small does it get on disk":

```{r footprint}
sizes <- sapply(fits, function(f) {
  tf <- tempfile(fileext = ".rds")
  saveRDS(f, tf, compress = "xz")
  on.exit(unlink(tf), add = TRUE)
  file.size(tf)
})

original <- as.numeric(object.size(draws))
data.frame(
  method        = names(sizes),
  bytes         = sizes,
  ratio_to_raw  = round(original / sizes, 1)
)
```

## Sample regeneration

All methods support [sample_posterior()] with the same interface:

```{r regenerate}
regen <- lapply(fits, sample_posterior, n_draws = 2000)

sapply(regen, function(s) {
  c(mean_alpha = mean(s[, "alpha"]),
    mean_beta  = mean(s[, "beta"]),
    mean_sigma = mean(s[, "sigma"]),
    cor_alpha_beta = cor(s[, "alpha"], s[, "beta"]))
})
```

## Pairwise comparison plot

Visual overlap between original draws (gray) and regenerated samples.

```{r combined-plot, fig.width=10, fig.height=8}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("patchwork", quietly = TRUE)) {
  draws_df <- as.data.frame(draws)
  palette_map <- c(mvdens_gmm = "red", mvdens_kde = "darkgreen", mclust = "blue")
  has_ggside <- requireNamespace("ggside", quietly = TRUE)

  plot_pairwise <- function(var1, var2) {
    p <- ggplot2::ggplot(mapping = ggplot2::aes_string(x = var1, y = var2)) +
      ggplot2::geom_point(
        data = draws_df,
        alpha = 0.15,
        size = 0.5,
        color = "gray50"
      ) +
      ggplot2::labs(
        x = var1,
        y = var2,
        title = paste(var1, "vs", var2)
      ) +
      ggplot2::theme_bw() +
      ggplot2::theme(plot.title = ggplot2::element_text(size = 9))

    if (has_ggside) {
      p <- p +
        ggside::geom_xsidedensity(
          data = draws_df,
          ggplot2::aes(y = ..density..),
          color = "gray50",
          alpha = 0.8
        ) +
        ggside::geom_ysidedensity(
          data = draws_df,
          ggplot2::aes(x = ..density..),
          color = "gray50",
          alpha = 0.8
        ) +
        ggside::theme_ggside_void()
    }

    for (method_name in names(regen)) {
      regen_df <- as.data.frame(regen[[method_name]])
      p <- p + ggplot2::geom_point(
        data = regen_df,
        alpha = 0.4,
        size = 0.5,
        color = unname(palette_map[method_name])
      )

      if (has_ggside) {
        p <- p +
          ggside::geom_xsidedensity(
            data = regen_df,
            ggplot2::aes(y = ..density..),
            color = unname(palette_map[method_name]),
            alpha = 0.8
          ) +
          ggside::geom_ysidedensity(
            data = regen_df,
            ggplot2::aes(x = ..density..),
            color = unname(palette_map[method_name]),
            alpha = 0.8
          )
      }
    }
    p
  }

  param_names <- colnames(draws_df)
  plot_list <- list()
  idx <- 1
  for (i in 1:(length(param_names) - 1)) {
    for (j in (i + 1):length(param_names)) {
      plot_list[[idx]] <- plot_pairwise(param_names[i], param_names[j])
      idx <- idx + 1
    }
  }

  ncol_grid <- ceiling(sqrt(length(plot_list)))
  combined_plot <- patchwork::wrap_plots(plot_list, ncol = ncol_grid) +
    patchwork::plot_annotation(
      title = "Original (gray) | mvdens GMM (red) | mvdens KDE (green) | mclust GMM (blue)"
    )

  combined_plot
} else {
  message("Install suggested packages 'ggplot2' and 'patchwork' to render combined_plot.")
}
```

## Density evaluation

```{r density}
test_pts <- rbind(c(5.5, 2.0, 2.0), c(6.2, 1.9, 1.8), c(7.0, 1.8, 2.1))

sapply(fits, function(f) density_posterior(f, test_pts))
```

## Picking a method

- `"mclust"`: fastest path; great default for parameter counts in the
  hundreds to low thousands.
- `"mvdens_gmm"`: alternative EM-based GMM, useful for cross-checks.
- `"mvdens_kde"`: non-parametric, larger storage footprint, useful when
  the posterior is far from a mixture of Gaussians.

## Session info

```{r si}
sessionInfo()
```
