---
title: "Comparing compression methods"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Comparing compression methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

`poco` exposes a single `method` argument to switch between density
approximations. This vignette compares the available backends on a
small synthetic posterior so you can pick the right trade-off for your
problem.

```{r libs}
library(poco)
library(mvtnorm)
```

## Available methods

```{r methods}
compression_methods()
```

| `method`        | Backend                    | Notes |
| --------------- | -------------------------- | ----- |
| `"mclust"`      | `mclust::Mclust()`         | parametric GMM, BIC-based selection, fast & compact |
| `"mvdens_gmm"`  | `mvdens::fit.gmm()`        | EM-based GMM, requires the suggested `mvdens` package |
| `"mvdens_kde"`  | `mvdens::fit.kde()`        | kernel density estimate, larger but non-parametric |

## A toy posterior with correlated parameters

```{r toy}
set.seed(1)
n_per <- 2000
draws <- rbind(
  mvtnorm::rmvnorm(n_per, mean = c(-2, -1),
                   sigma = matrix(c(0.5,  0.1,  0.1, 0.3), 2)),
  mvtnorm::rmvnorm(n_per, mean = c( 2,  1),
                   sigma = matrix(c(0.4, -0.2, -0.2, 0.6), 2))
)
colnames(draws) <- c("alpha", "beta")
```

## Compress with each method

`mvdens_*` methods require the suggested `mvdens` package; we skip them
gracefully when it is not installed.

```{r compress}
methods_to_try <- c("mclust")
if (requireNamespace("mvdens", quietly = TRUE)) {
  methods_to_try <- c(methods_to_try, "mvdens_gmm", "mvdens_kde")
}

fits <- lapply(methods_to_try, function(m) {
  compress_posterior(draws, method = m, n_components = 2)
})
names(fits) <- methods_to_try
```

## Storage footprint

A practical proxy for "how small does it get on disk":

```{r footprint}
sizes <- sapply(fits, function(f) {
  tf <- tempfile(fileext = ".rds")
  saveRDS(f, tf, compress = "xz")
  on.exit(unlink(tf), add = TRUE)
  file.size(tf)
})

original <- as.numeric(object.size(draws))
data.frame(
  method        = names(sizes),
  bytes         = sizes,
  ratio_to_raw  = round(original / sizes, 1)
)
```

## Sample regeneration

All methods support [sample_posterior()] with the same interface:

```{r regenerate}
regen <- lapply(fits, sample_posterior, n_draws = 2000)

sapply(regen, function(s) {
  c(mean_alpha = mean(s[, "alpha"]),
    mean_beta  = mean(s[, "beta"]),
    cor        = cor(s[, "alpha"], s[, "beta"]))
})
```

## Density evaluation

```{r density}
test_pts <- rbind(c(-2, -1), c(0, 0), c(2, 1))

sapply(fits, function(f) density_posterior(f, test_pts))
```

## Picking a method

- `"mclust"`: fastest path; great default for parameter counts in the
  hundreds to low thousands.
- `"mvdens_gmm"`: alternative EM-based GMM, useful for cross-checks.
- `"mvdens_kde"`: non-parametric, larger storage footprint, useful when
  the posterior is far from a mixture of Gaussians.

## Session info

```{r si}
sessionInfo()
```
