---
title: "Tuning `compress_brmsfit` on posteriorHCA models"
subtitle: "Benchmark report from the multi cell-type, multi-gene targets pipeline"
date: today
date-format: "DD MMM YYYY"
title-block-banner: true
author: 
  - name: Chen Zhan
    orcid: 0000-0002-4794-8339
    email: chen.zhan@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
format: 
  html:
    theme: minty
    df-print: paged
    code-line-numbers: true
    embed-resources: true
    code-fold: true
    code-fold-default: true
    toc: true
    toc-location: left
    toc-title: "Contents"
knitr:
  opts_chunk: 
    message: false
    warning: false
    fig.path: ../../inst/figures/brms-parameter-tuning/
    fig.ext: .png
    dev: png
    dpi: 150
editor: source
abstract: |
  We compressed `brms` posteriors from the posteriorHCA differential-expression
  workflow with **poco**, sweeping `n_components` and `corr_threshold` across
  many cell types and genes. Fitting and evaluation were run in a separate
  [`targets`](https://books.ropensci.org/targets/) pipeline; this document
  summarises the saved grid (`evaluation_results_targets_V2.rds`): where
  compression fails, how long it takes, which parameter pair works best on
  average, whether cell types agree, and whether gene abundance shifts the
  optimum.
---

## Setup

```{r}
#| label: setup

# Figures are written as PNG under poco/inst/figures/brms-parameter-tuning/
fig_dir <- "../../inst/figures/brms-parameter-tuning"
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(glue)
library(forcats)
library(stringr)
library(patchwork)
library(ggrepel)

```

## Reproducing the benchmark (optional)

The grid was built with `posteriorHCA_brms_fit_compression_parameter_tunning_pipeline_V2.R` and `{targets}`. Re-run only when you need fresh fits:

```{r}
#| eval: false

source("posteriorHCA_brms_fit_compression_parameter_tunning_pipeline_V2.R")
targets::tar_make()
```

```{r}
#| eval: false
#| 
compression_results <-
  tar_meta(starts_with("compression_"), store = "_targets") %>%
  filter(type == "branch") %>%
  slice_sample(n = 1) %>% 
  pull(name) %>%
  map_dfr(~ {
    t <- tar_read_raw(.x, store = "_targets")
    
    if (isTRUE(t$compression_ok)) {
      
      o <- tar_read_raw(
      t$estimates_chunk_names, 
      store = file.path(
        '/hpcfs/groups/phoenix-hpc-mangiola_laboratory/chen/HPC_posterior/NEW_CELL_TYPE/', 
        t$cell_type, 
        '_targets'
      )
    )
      
      t$compression_size <- as.numeric(
        lobstr::obj_size(t$compressed_obj[[1]]$compressed)
      )
      t$original_draw_size <- as.numeric(
        lobstr::obj_size(o$brms_fit[[1]] %>% as_draws_matrix)
      )
      t$compression_rate <- t$compression_size / t$original_draw_size
      
    } else {
      t$compression_size <- NA_real_
      t$original_draw_size <- NA_real_
      t$compression_rate <- NA_real_
    }

    t %>%
      dplyr::select(-compressed_obj)
  })

```

## Load results

```{r}
#| label: load-results

param_tunning_long <- readRDS("evaluation_results_targets_V2.rds")
compression_results_long <- readRDS('compression_size.rds')
```

## Prepare the tuning table

Each row is one `(cell_type, gene, n_components, corr_threshold)` combination.
Pseudobulk abundance (`mean_log2_cpm`) is included so we can relate the optimum
to expression level later.

```{r}
#| label: tidy-param-tunning

n_genes_per_ct <- param_tunning_long %>%
  distinct(cell_type, .feature) %>%
  count(cell_type, name = "n_genes")

cat("Cell types and genes in the tuning grid:\n")
print(n_genes_per_ct)
cat("\nGrid: ",
    n_distinct(param_tunning_long$n_components), " n_components x ",
    n_distinct(param_tunning_long$corr_threshold), " corr_threshold values\n",
    sep = "")
```

# 1. Compression failures

Rows with `reproduction_pct == NA` did not receive a fidelity score—typically
because the `brms` fit could not be read or `compress_brmsfit()` /
`evaluate_compression()` failed for that grid cell. Mapping these gaps shows
which parameter combinations, cell types, or genes to exclude before choosing
defaults.

```{r}
#| label: failure-flag

param_tunning_long <-
  param_tunning_long %>%
  mutate(failed = is.na(reproduction_pct))

failure_overall <- param_tunning_long %>%
  summarise(
    n_total    = n(),
    n_failed   = sum(failed),
    pct_failed = 100 * mean(failed)
  )

cat(sprintf(
  "Overall: %d / %d grid rows have reproduction_pct == NA (%.2f%%)\n",
  failure_overall$n_failed,
  failure_overall$n_total,
  failure_overall$pct_failed
))
```

When failure rates are low, treat the plots below as sanity checks; when they
are high, use them to decide what to filter.

## 1.1 Failure rate on the parameter grid

A hot grid cell that fails for many genes and cell types usually indicates a
parameter setting the compressor cannot handle (for example, too many mixture
components for small correlation blocks), rather than a biological effect.

```{r}
#| label: failure-grid-heatmap
#| fig-width: 9
#| fig-height: 5

failure_grid <-
  param_tunning_long %>%
  group_by(n_components, corr_threshold) %>%
  summarise(
    n_total    = n(),
    n_failed   = sum(failed),
    pct_failed = 100 * mean(failed),
    .groups    = "drop"
  )

ggplot(
  failure_grid,
  aes(
    x    = factor(corr_threshold),
    y    = factor(n_components),
    fill = pct_failed
  )
) +
  geom_tile(colour = "grey92", linewidth = 0.35) +
  geom_text(
    aes(label = sprintf("%.0f%%\n(%d/%d)",
                        pct_failed, n_failed, n_total)),
    size = 2.8, lineheight = 0.9
  ) +
  scale_fill_gradient(
    name     = "% failed",
    low      = "white",
    high     = "#b2182b",
    limits   = c(0, max(100, max(failure_grid$pct_failed, na.rm = TRUE))),
    na.value = "grey85"
  ) +
  labs(
    title    = "Compression failure rate on the tuning grid",
    subtitle = "% of (cell type, gene) pairs with missing reproduction_pct",
    x = "corr_threshold (partition cut height)",
    y = "n_components (mclust per cluster)"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())
```

## 1.2 Failure rate by cell type

```{r}
#| label: failure-by-celltype
#| fig-width: 9
#| fig-height: 5

failure_per_ct <-
  param_tunning_long %>%
  group_by(cell_type) %>%
  summarise(
    n_total    = n(),
    n_failed   = sum(failed),
    pct_failed = 100 * mean(failed),
    n_genes    = n_distinct(.feature),
    n_genes_any_fail = n_distinct(.feature[failed]),
    n_genes_all_fail = sum(
      tapply(failed, .feature, function(x) all(x))
    ),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_failed))

cat("Failure summary per cell type:\n")
print(failure_per_ct)

ggplot(
  failure_per_ct,
  aes(x = fct_reorder(cell_type, pct_failed), y = pct_failed)
) +
  geom_col(fill = "#b2182b", alpha = 0.85) +
  geom_text(
    aes(label = sprintf("%d/%d (%.0f%%)", n_failed, n_total, pct_failed)),
    hjust = -0.05, size = 3
  ) +
  coord_flip() +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15)),
    labels = function(x) paste0(x, "%")
  ) +
  labs(
    title = "Compression failure rate per cell type",
    subtitle = "Aggregated over the full (n_components, corr_threshold) grid",
    x = NULL,
    y = "% of grid rows failed"
  ) +
  theme_bw(base_size = 12)
```

## 1.3 Failure surface by cell type

Same view as Section 1.1, faceted by cell type—useful for seeing whether a bad
parameter cell is universal or specific to one lineage.

```{r}
#| label: failure-grid-by-celltype
#| fig-width: 12
#| fig-height: 8

failure_grid_ct <-
  param_tunning_long %>%
  group_by(cell_type, n_components, corr_threshold) %>%
  summarise(
    n_total    = n(),
    n_failed   = sum(failed),
    pct_failed = 100 * mean(failed),
    .groups    = "drop"
  )

ggplot(
  failure_grid_ct,
  aes(
    x    = factor(corr_threshold),
    y    = factor(n_components),
    fill = pct_failed
  )
) +
  geom_tile(colour = "grey95", linewidth = 0.2) +
  facet_wrap(~ cell_type, ncol = 5) +
  scale_fill_gradient(
    name     = "% failed",
    low      = "white",
    high     = "#b2182b",
    limits   = c(0, 100),
    na.value = "grey85"
  ) +
  labs(
    title    = "Failure surface per cell type",
    subtitle = "Empty / white = no failures, dark = all genes failed at that grid cell",
    x = "corr_threshold",
    y = "n_components"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid  = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    strip.text  = element_text(size = 9)
  )
```

## 1.4 Failures vs gene abundance

If failures cluster at low or high expression, a single universal default
(Section 5) may only work for part of the abundance range.

```{r}
#| label: failure-by-gene-abundance
#| fig-width: 10
#| fig-height: 5

gene_failure <-
  param_tunning_long %>%
  group_by(cell_type, .feature, mean_log2_cpm) %>%
  summarise(
    n_total    = n(),
    n_failed   = sum(failed),
    pct_failed = 100 * mean(failed),
    .groups    = "drop"
  )

ggplot(
  gene_failure %>% filter(!is.na(mean_log2_cpm)),
  aes(x = mean_log2_cpm, y = pct_failed)
) +
  geom_point(aes(colour = cell_type), size = 1.8, alpha = 0.7) +
  geom_smooth(method = "loess", se = TRUE, colour = "grey20") +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     limits = c(0, 100)) +
  scale_colour_viridis_d(option = "turbo", name = "cell_type") +
  labs(
    title    = "Per-gene compression failure fraction vs abundance",
    subtitle = "y = % of the (n_components, corr_threshold) grid that failed for this gene",
    x = "gene mean log2 CPM (pseudobulk)",
    y = "% of grid failed"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "right",
        legend.text     = element_text(size = 7),
        legend.key.size = unit(0.4, "cm"))
```

## 1.5 Failures vs maximum block size

`block_size_max` is the largest cluster in the correlation partition before
compression. Larger blocks imply higher-dimensional mixture fits and may
coincide with more failures.

```{r}
#| label: failure-vs-block-size-max
#| fig-width: 11
#| fig-height: 5.5

blk <- param_tunning_long %>%
  filter(!is.na(block_size_max))

sp <- suppressWarnings(
  cor.test(blk$block_size_max, as.numeric(blk$failed), method = "spearman")
)
cat(sprintf(
  "Spearman rho (block_size_max vs failed): %.3f, p = %.3g\n",
  unname(sp$estimate), sp$p.value
))

p_box <- ggplot(
  blk,
  aes(
    x = factor(failed, levels = c(FALSE, TRUE), labels = c("OK", "failed")),
    y = block_size_max,
    fill = failed
  )
) +
  geom_boxplot(outlier.alpha = 0.06, linewidth = 0.35) +
  scale_y_continuous(trans = "sqrt", labels = label_number()) +
  scale_fill_manual(values = c(`FALSE` = "#4393c3", `TRUE` = "#b2182b"), guide = "none") +
  labs(
    title = "Maximum block size by outcome",
    x = NULL,
    y = "block_size_max (sqrt scale)"
  ) +
  theme_bw(base_size = 11)

blk_bin <- blk %>%
  mutate(bmax_grp = cut_number(block_size_max, n = 20)) %>%
  group_by(bmax_grp) %>%
  summarise(
    n = n(),
    pct_fail = 100 * mean(failed),
    m_mid = median(block_size_max),
    .groups = "drop"
  )

p_rate <- ggplot(blk_bin, aes(x = m_mid, y = pct_fail, size = n)) +
  geom_point(alpha = 0.8, colour = "#7f1d1d") +
  geom_smooth(aes(weight = n), method = "loess", se = TRUE, linewidth = 0.65, colour = "grey25") +
  scale_size_area(max_size = 7, name = "n rows") +
  scale_y_continuous(limits = c(0, NA), labels = function(x) paste0(x, "%")) +
  labs(
    title = "Failure rate along block_size_max",
    x = "median block_size_max in bin",
    y = "% failed"
  ) +
  theme_bw(base_size = 11)

p_box + p_rate + plot_layout(ncol = 2)
```

## 1.6 Gene-wide vs grid-wide failures

- **Whole-gene failure** — every grid cell failed: likely a bad `brms` fit; drop the gene.
- **Whole-grid-cell failure** — every gene failed at one `(n_components, corr_threshold)`: drop that cell from the universal search.

```{r}
#| label: failure-decomposition

gene_all_failed <-
  param_tunning_long %>%
  group_by(cell_type, .feature) %>%
  summarise(all_failed = all(failed), any_failed = any(failed),
            .groups = "drop") %>%
  summarise(
    n_genes              = n(),
    n_genes_any_failure  = sum(any_failed),
    n_genes_all_failure  = sum(all_failed),
    pct_genes_all_failure = 100 * mean(all_failed)
  )

grid_all_failed <-
  param_tunning_long %>%
  group_by(n_components, corr_threshold) %>%
  summarise(all_failed = all(failed), .groups = "drop") %>%
  summarise(
    n_grid_cells          = n(),
    n_grid_cells_all_fail = sum(all_failed),
    pct_grid_cells_all_fail = 100 * mean(all_failed)
  )

cat("Whole-gene failures (every grid cell NA for that gene):\n")
print(gene_all_failed)

cat("\nWhole-grid-cell failures (every gene NA at that (n_comp, thr)):\n")
print(grid_all_failed)
```

### 1.6.1 Failure pattern per gene on the grid

For each cell type we plot up to 25 whole-gene failures plus 20 genes with
the highest partial failure rate. Light tiles succeeded; dark tiles failed.

```{r}
#| label: failure-gene-grid-heatmap
#| fig-width: 14
#| fig-height: 10

gene_fail_stats <- param_tunning_long %>%
  group_by(cell_type, .feature) %>%
  summarise(
    pct_failed = 100 * mean(failed),
    whole_gene_fail = all(failed),
    .groups = "drop"
  )

genes_for_grid <- gene_fail_stats %>%
  filter(pct_failed > 0) %>%
  group_by(cell_type) %>%
  group_modify(~ bind_rows(
    .x %>% filter(whole_gene_fail) %>% slice_head(n = 25),
    .x %>% filter(!whole_gene_fail) %>% slice_max(pct_failed, n = 20)
  )) %>%
  distinct(cell_type, .feature) %>%
  ungroup()

fail_tile <- param_tunning_long %>%
  semi_join(genes_for_grid, by = c("cell_type", ".feature")) %>%
  mutate(
    grid_id = interaction(factor(n_components), factor(corr_threshold), sep = " | "),
    gene_lab = str_trunc(as.character(.feature), 22),
    status = if_else(failed, "failed", "OK")
  ) %>%
  left_join(
    gene_fail_stats %>%
      mutate(
        gene_lab = str_trunc(as.character(.feature), 22),
        ord = as.numeric(whole_gene_fail) * 1e6 + pct_failed
      ) %>%
      select(cell_type, gene_lab, ord),
    by = c("cell_type", "gene_lab")
  ) %>%
  group_by(cell_type) %>%
  mutate(row_lab_f = fct_reorder(factor(gene_lab), ord, .fun = mean)) %>%
  ungroup()

ncol_ct <- min(4L, n_distinct(fail_tile$cell_type))

ggplot(fail_tile, aes(grid_id, row_lab_f, fill = status)) +
  geom_tile(colour = "grey88", linewidth = 0.15) +
  scale_x_discrete(guide = guide_axis(angle = 55)) +
  scale_fill_manual(values = c(OK = "grey96", failed = "#b2182b")) +
  facet_wrap(~ cell_type, ncol = ncol_ct, scales = "free_y") +
  labs(
    title = "Where each gene fails on the tuning grid",
    x = "n_components | corr_threshold",
    y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank(), strip.text = element_text(face = "bold", size = 9))
```

High `pct_genes_all_failure` means later global summaries use fewer genes.
Drop any `(n_components, corr_threshold)` with `grid_cells_all_fail` before
Section 3. Grid means use `na.rm = TRUE`; per-gene bests use rows with a
valid `reproduction_pct`.

::: {.callout-important}
## Why compression failed (and why this is no longer a concern)

The missing `reproduction_pct` values above come from two distinct causes:

1. **BiocParallel / `SnowParam` port errors** (e.g. *cannot find an open
   port*) — almost certainly an artefact of the HPC execution environment,
   not of **poco** itself. They are uncommon, typically cleared by
   re-running the affected `{targets}` step, and should not be interpreted
   as a systematic compression defect.

2. **Over-parameterised `mclust` fits during tuning** — the remaining
   failures reflect a mismatch between the mixture complexity requested on
   the grid and the effective dimensionality of the posterior draws: many
   parameters per correlation block, but only a limited number of MCMC
   draws to support them. We originally required correlated blocks to use
   **full-covariance** mixtures so pairwise structure can be preserved; at
   extreme `(n_components, corr_threshold)` settings that model could not
   be estimated.

The current **poco** release still prefers full-covariance mixtures for
correlated clusters, but **automatically switches to a simpler model when a
fit fails**. Compression failures of the kind plotted in this section are
therefore resolved in the updated package and are not expected to affect
the parameter choices in Sections 3–6.
:::

# 2. Compression efficiency (storage footprint)

`compression_rate` is `compression_size / original_draw_size` (RAM footprint of
the compressed draws matrix vs the uncompressed `as_draws_matrix` object).
Smaller values mean a tighter representation. Summaries use rows with
`compression_ok == TRUE` and non-missing `compression_rate`, loaded from
`compression_size.rds` (`compression_results_long`).

```{r}
#| label: compression-rate-overview

comp_rate_ok <- compression_results_long %>%
  filter(as.logical(compression_ok), !is.na(compression_rate))

cat(sprintf(
  "Compression size section: %d / %d rows with successful compression and non-NA compression_rate.\n\n",
  nrow(comp_rate_ok),
  nrow(compression_results_long)
))

rate_overall <- comp_rate_ok %>%
  summarise(
    n_rows      = n(),
    min_rate    = min(compression_rate, na.rm = TRUE),
    q25_rate    = quantile(compression_rate, 0.25, na.rm = TRUE),
    median_rate = median(compression_rate, na.rm = TRUE),
    mean_rate   = mean(compression_rate, na.rm = TRUE),
    q75_rate    = quantile(compression_rate, 0.75, na.rm = TRUE),
    max_rate    = max(compression_rate, na.rm = TRUE)
  )

cat("Overall compression_rate (compressed / original), successful runs only:\n")
print(rate_overall)

rate_grid <- comp_rate_ok %>%
  group_by(n_components, corr_threshold) %>%
  summarise(
    n           = n(),
    median_rate = median(compression_rate, na.rm = TRUE),
    mean_rate   = mean(compression_rate, na.rm = TRUE),
    q25_rate    = quantile(compression_rate, 0.25, na.rm = TRUE),
    q75_rate    = quantile(compression_rate, 0.75, na.rm = TRUE),
    mean_blocks = mean(n_blocks, na.rm = TRUE),
    .groups     = "drop"
  ) %>%
  arrange(median_rate)

cat("\nTightest grid cells by median compression_rate (top 12), successful runs only:\n")
print(head(rate_grid, 12))
```

### 2.1 Median compression rate on the parameter grid

```{r}
#| label: compression-rate-grid-heatmap
#| fig-width: 9
#| fig-height: 5

rate_heat <- comp_rate_ok %>%
  group_by(n_components, corr_threshold) %>%
  summarise(median_rate = median(compression_rate, na.rm = TRUE), .groups = "drop")

ggplot(
  rate_heat,
  aes(
    x    = factor(corr_threshold),
    y    = factor(n_components),
    fill = median_rate
  )
) +
  geom_tile(colour = "grey92", linewidth = 0.35) +
  geom_text(aes(label = scales::percent(median_rate, accuracy = 0.1)), size = 2.8) +
  scale_fill_viridis_c(
    option    = "C",
    name      = "median\nrate",
    labels    = scales::percent_format(accuracy = 1),
    direction = -1,
    na.value  = "grey85"
  ) +
  labs(
    title    = "Median compression rate across the tuning grid",
    subtitle = "Successful compressions only; lower = smaller footprint vs uncompressed draws",
    x        = "corr_threshold",
    y        = "n_components"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())
```

### 2.2 Compression rate distributions by tuning parameters

```{r}
#| label: compression-rate-param-boxplots
#| fig-width: 11
#| fig-height: 5

p_rate_nc <- ggplot(
  comp_rate_ok %>% filter(!is.na(compression_rate)),
  aes(x = factor(n_components), y = compression_rate)
) +
  geom_boxplot(outlier.alpha = 0.15, fill = "#4393c3", colour = "grey30") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Compression rate by n_components",
    subtitle = "Successful compressions only; rate = compressed size / original draws size",
    x     = "n_components",
    y     = "compression_rate"
  ) +
  theme_bw(base_size = 11)

p_rate_corr <- ggplot(
  comp_rate_ok %>% filter(!is.na(compression_rate)),
  aes(x = factor(corr_threshold), y = compression_rate)
) +
  geom_boxplot(outlier.alpha = 0.15, fill = "#d6604d", colour = "grey30") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Compression rate by corr_threshold",
    subtitle = "Successful compressions only",
    x     = "corr_threshold",
    y     = "compression_rate"
  ) +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_rate_nc + p_rate_corr + plot_layout(ncol = 2)
```

### 2.3 Compression rate vs partition size and abundance

Finer partitions (more blocks) or genes at different abundance may shift how much
structure the mixture must retain, which feeds through to object size.

```{r}
#| label: compression-rate-covariates
#| fig-width: 11
#| fig-height: 5.5

dat_rate_blocks <- comp_rate_ok %>%
  filter(!is.na(compression_rate), !is.na(n_blocks))

pr_blocks <- ggplot(dat_rate_blocks, aes(x = n_blocks, y = compression_rate, colour = factor(n_components))) +
  geom_point(alpha = 0.12, size = 0.6) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.9, colour = "grey25") +
  scale_colour_viridis_d(option = "turbo", name = "n_components") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "compression_rate vs n_blocks",
    subtitle = "Successful compressions only; colour = n_components",
    x     = "n_blocks (partition)",
    y     = "compression_rate"
  ) +
  theme_bw(base_size = 11)

dat_rate_abund <- comp_rate_ok %>%
  filter(!is.na(compression_rate), !is.na(mean_log2_cpm))

pr_abund <- ggplot(dat_rate_abund, aes(x = mean_log2_cpm, y = compression_rate)) +
  geom_bin2d(bins = 40) +
  scale_fill_viridis_c(option = "B", end = 0.9, name = "count") +
  geom_smooth(method = "loess", se = TRUE, colour = "orangered", linewidth = 0.9) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "compression_rate vs mean_log2_cpm",
    subtitle = "Successful compressions only",
    x     = "mean log2 CPM",
    y     = "compression_rate"
  ) +
  theme_bw(base_size = 11)

pr_blocks + pr_abund + plot_layout(ncol = 2)
```

```{r}
#| label: compression-rate-spearman

rate_cor <- comp_rate_ok %>%
  filter(!is.na(compression_rate)) %>%
  summarise(
    cor_rate_blocks = suppressWarnings(
      cor(compression_rate, n_blocks, use = "pairwise.complete.obs", method = "spearman")
    ),
    cor_rate_bmax = suppressWarnings(
      cor(compression_rate, block_size_max, use = "pairwise.complete.obs", method = "spearman")
    ),
    cor_rate_bmed = suppressWarnings(
      cor(compression_rate, block_size_med, use = "pairwise.complete.obs", method = "spearman")
    ),
    cor_rate_abund = suppressWarnings(
      cor(compression_rate, mean_log2_cpm, use = "pairwise.complete.obs", method = "spearman")
    )
  )

cat("Spearman correlation of compression_rate with covariates (successful compressions only):\n")
print(rate_cor)
```

### 2.4 Median compression rate by cell type

```{r}
#| label: compression-rate-by-celltype
#| fig-width: 9
#| fig-height: 5.5

rate_ct <- comp_rate_ok %>%
  group_by(cell_type) %>%
  summarise(
    n           = n(),
    median_rate = median(compression_rate, na.rm = TRUE),
    mean_rate   = mean(compression_rate, na.rm = TRUE),
    q75_rate    = quantile(compression_rate, 0.75, na.rm = TRUE),
    .groups     = "drop"
  ) %>%
  arrange(median_rate)

cat("Median compression_rate per cell type (successful compressions only):\n")
print(rate_ct)

ggplot(rate_ct, aes(x = fct_reorder(cell_type, median_rate), y = median_rate)) +
  geom_col(fill = "#2166ac", alpha = 0.9) +
  geom_errorbar(
    aes(ymin = median_rate, ymax = q75_rate),
    width = 0.2,
    colour = "grey35"
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    title    = "Median compression rate per cell type",
    subtitle = "Successful runs; error bar top = 75th percentile of compression_rate within type",
    x        = NULL,
    y        = "median compression_rate"
  ) +
  theme_bw(base_size = 12)
```

# 3. Compression runtime

`time_used` is seconds for `compress_brmsfit()` only (partitioning excluded).
Summaries below use rows with `compression_ok == TRUE`.

```{r}
#| label: time-used-overview

param_time_ok <- param_tunning_long %>%
  filter(as.logical(compression_ok))

cat(sprintf(
  "Time section: %d / %d rows with successful compression (compression_ok is true).\n\n",
  nrow(param_time_ok),
  nrow(param_tunning_long)
))

time_overall <- param_time_ok %>%
  summarise(
    n_rows      = n(),
    n_na_time   = sum(is.na(time_used)),
    min_s       = min(time_used, na.rm = TRUE),
    q25_s       = quantile(time_used, 0.25, na.rm = TRUE),
    median_s    = median(time_used, na.rm = TRUE),
    mean_s      = mean(time_used, na.rm = TRUE),
    q75_s       = quantile(time_used, 0.75, na.rm = TRUE),
    max_s       = max(time_used, na.rm = TRUE)
  )

cat("Overall time_used (seconds), successful compressions only:\n")
print(time_overall)

time_grid <- param_time_ok %>%
  group_by(n_components, corr_threshold) %>%
  summarise(
    n          = n(),
    median_s   = median(time_used, na.rm = TRUE),
    mean_s     = mean(time_used, na.rm = TRUE),
    q25_s      = quantile(time_used, 0.25, na.rm = TRUE),
    q75_s      = quantile(time_used, 0.75, na.rm = TRUE),
    mean_blocks = mean(n_blocks, na.rm = TRUE),
    .groups    = "drop"
  ) %>%
  arrange(desc(median_s))

cat("\nSlowest grid cells by median time_used (top 12), successful runs only:\n")
print(head(time_grid, 12))
```

## 3.1 Median runtime on the parameter grid

```{r}
#| label: time-used-grid-heatmap
#| fig-width: 9
#| fig-height: 5

time_heat <- param_time_ok %>%
  group_by(n_components, corr_threshold) %>%
  summarise(median_s = median(time_used, na.rm = TRUE), .groups = "drop")

ggplot(
  time_heat,
  aes(
    x    = factor(corr_threshold),
    y    = factor(n_components),
    fill = median_s
  )
) +
  geom_tile(colour = "grey92", linewidth = 0.35) +
  geom_text(aes(label = sprintf("%.1fs", median_s)), size = 2.8) +
  scale_fill_viridis_c(
    option = "C",
    name   = "median\nseconds",
    trans  = "sqrt",
    labels = label_number(accuracy = 0.1)
  ) +
  labs(
    title    = "Median compression time across the tuning grid",
    subtitle = "Successful compressions only; sqrt colour scale for fast vs slow cells",
    x        = "corr_threshold",
    y        = "n_components"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())
```

## 3.2 Runtime distributions

```{r}
#| label: time-used-param-boxplots
#| fig-width: 11
#| fig-height: 5

p_nc <- ggplot(
  param_time_ok %>% filter(!is.na(time_used)),
  aes(x = factor(n_components), y = time_used)
) +
  geom_boxplot(outlier.alpha = 0.15, fill = "#4393c3", colour = "grey30") +
  scale_y_continuous(labels = label_number(suffix = " s"), trans = "sqrt") +
  labs(
    title = "Time by n_components",
    subtitle = "Successful compressions only; sqrt y-axis for long tail",
    x     = "n_components",
    y     = "time_used (s)"
  ) +
  theme_bw(base_size = 11)

p_corr <- ggplot(
  param_time_ok %>% filter(!is.na(time_used)),
  aes(x = factor(corr_threshold), y = time_used)
) +
  geom_boxplot(outlier.alpha = 0.15, fill = "#d6604d", colour = "grey30") +
  scale_y_continuous(labels = label_number(suffix = " s"), trans = "sqrt") +
  labs(
    title = "Time by corr_threshold",
    subtitle = "Successful compressions only",
    x     = "corr_threshold",
    y     = "time_used (s)"
  ) +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_nc + p_corr + plot_layout(ncol = 2)
```

## 3.3 Runtime vs partition size and abundance

More blocks or larger blocks usually mean more mixture fits inside
`compress_brmsfit()`, so wall time should rise with partition complexity.

```{r}
#| label: time-used-covariates
#| fig-width: 11
#| fig-height: 5.5

dat_blocks <- param_time_ok %>%
  filter(!is.na(time_used), !is.na(n_blocks))

pu_blocks <- ggplot(dat_blocks, aes(x = n_blocks, y = time_used, colour = factor(n_components))) +
  geom_point(alpha = 0.12, size = 0.6) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.9, colour = "grey25") +
  scale_colour_viridis_d(option = "turbo", name = "n_components") +
  scale_y_continuous(labels = label_number(suffix = " s"), trans = "sqrt") +
  labs(
    title = "time_used vs n_blocks",
    subtitle = "Successful compressions only; colour = n_components",
    x     = "n_blocks (partition)",
    y     = "time_used (s)"
  ) +
  theme_bw(base_size = 11)

dat_abund <- param_time_ok %>%
  filter(!is.na(time_used), !is.na(mean_log2_cpm))

pu_abund <- ggplot(dat_abund, aes(x = mean_log2_cpm, y = time_used)) +
  geom_bin2d(bins = 40) +
  scale_fill_viridis_c(option = "B", end = 0.9, name = "count") +
  geom_smooth(method = "loess", se = TRUE, colour = "orangered", linewidth = 0.9) +
  scale_y_continuous(labels = label_number(suffix = " s"), trans = "sqrt") +
  labs(
    title = "time_used vs mean_log2_cpm",
    subtitle = "Successful compressions only",
    x     = "mean log2 CPM",
    y     = "time_used (s)"
  ) +
  theme_bw(base_size = 11)

pu_blocks + pu_abund + plot_layout(ncol = 2)
```

```{r}
#| label: time-used-spearman

time_cor <- param_time_ok %>%
  filter(!is.na(time_used)) %>%
  summarise(
    cor_time_blocks = suppressWarnings(
      cor(time_used, n_blocks, use = "pairwise.complete.obs", method = "spearman")
    ),
    cor_time_bmax = suppressWarnings(
      cor(time_used, block_size_max, use = "pairwise.complete.obs", method = "spearman")
    ),
    cor_time_bmed = suppressWarnings(
      cor(time_used, block_size_med, use = "pairwise.complete.obs", method = "spearman")
    ),
    cor_time_abund = suppressWarnings(
      cor(time_used, mean_log2_cpm, use = "pairwise.complete.obs", method = "spearman")
    )
  )

cat("Spearman correlation of time_used with covariates (successful compressions only):\n")
print(time_cor)
```

## 3.4 Median runtime by cell type

```{r}
#| label: time-used-by-celltype
#| fig-width: 9
#| fig-height: 5.5

time_ct <- param_time_ok %>%
  group_by(cell_type) %>%
  summarise(
    n        = n(),
    median_s = median(time_used, na.rm = TRUE),
    mean_s   = mean(time_used, na.rm = TRUE),
    q75_s    = quantile(time_used, 0.75, na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  arrange(desc(median_s))

cat("Median time_used (s) per cell type (successful compressions only):\n")
print(time_ct)

ggplot(time_ct, aes(x = fct_reorder(cell_type, median_s), y = median_s)) +
  geom_col(fill = "#2166ac", alpha = 0.9) +
  geom_errorbar(
    aes(ymin = median_s, ymax = q75_s),
    width = 0.2,
    colour = "grey35"
  ) +
  coord_flip() +
  scale_y_continuous(labels = label_number(suffix = " s"), expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "Median compression time per cell type",
    subtitle = "Successful runs; error bar top = 75th percentile of time_used within type",
    x        = NULL,
    y        = "median time_used (s)"
  ) +
  theme_bw(base_size = 12)
```

# 4. Universal compression setting

For each `(n_components, corr_threshold)` we average `reproduction_pct`
(mean of energy and C2ST fidelity) over all genes and cell types. The
heatmap shows the global surface; the circled cell is the best default,
and the dashed outline marks cells within 0.5 percentage points of that peak.

```{r}
#| label: overall-grid-summary

overall_grid <-
  param_tunning_long %>%
  group_by(n_components, corr_threshold) %>%
  summarise(
    mean_repro   = mean(reproduction_pct,   na.rm = TRUE),
    median_repro = median(reproduction_pct, na.rm = TRUE),
    sd_repro     = sd(reproduction_pct,     na.rm = TRUE),
    q25_repro    = quantile(reproduction_pct, 0.25, na.rm = TRUE),
    q75_repro    = quantile(reproduction_pct, 0.75, na.rm = TRUE),
    mean_blocks  = mean(n_blocks,           na.rm = TRUE),
    n_obs        = sum(!is.na(reproduction_pct)),
    .groups      = "drop"
  )

best_overall <- overall_grid %>%
  filter(!is.na(mean_repro)) %>%
  slice_max(mean_repro, n = 1, with_ties = FALSE, na_rm = TRUE)
plateau_thr  <- best_overall$mean_repro - 0.5

n_pairs_total <- nrow(distinct(param_tunning_long, cell_type, .feature))
n_pairs_ok    <- nrow(distinct(
  param_tunning_long %>% filter(!failed),
  cell_type, .feature
))

cat("Best overall (max mean reproduction_pct):\n")
print(best_overall)
```

## 4.1 Mean reproduction heatmap

```{r}
#| label: overall-heatmap
#| fig-width: 9
#| fig-height: 5

repro_rng    <- range(overall_grid$mean_repro, na.rm = TRUE)
repro_pad    <- max(0.25, 0.03 * diff(repro_rng))
repro_limits <- c(repro_rng[1] - repro_pad, repro_rng[2] + repro_pad)

ggplot(
  overall_grid,
  aes(
    x    = factor(corr_threshold),
    y    = factor(n_components),
    fill = mean_repro
  )
) +
  geom_tile(colour = "grey92", linewidth = 0.35) +
  geom_text(
    aes(label = sprintf("%.1f", mean_repro)),
    colour = "black", size = 3
  ) +
  # within-0.5pp plateau
  geom_tile(
    data = overall_grid %>% filter(mean_repro >= plateau_thr),
    fill = NA, colour = "grey20", linewidth = 0.6, linetype = "dashed"
  ) +
  # the single best cell
  geom_point(
    data = best_overall,
    inherit.aes = FALSE,
    aes(
      x = factor(corr_threshold),
      y = factor(n_components)
    ),
    shape = 21, size = 9, stroke = 1.2,
    fill = NA, colour = "black"
  ) +
  scale_fill_gradient(
    name = sprintf(
      "mean reproduction %%\n(range %.1f-%.1f)",
      repro_rng[1], repro_rng[2]
    ),
    low      = "white",
    high     = "#b2182b",
    limits   = repro_limits,
    oob      = scales::squish,
    na.value = "grey85"
  ) +
  labs(
    title    = "Universal compression-fidelity grid",
    subtitle = glue(
      "Means exclude NA; {n_pairs_ok} of {n_pairs_total} (cell_type, gene) pairs contribute | ",
      "best = n_components={best_overall$n_components}, corr_threshold={best_overall$corr_threshold} ",
      "({sprintf('%.2f%%', best_overall$mean_repro)})"
    ),
    x = "corr_threshold (partition cut height)",
    y = "n_components (mclust per cluster)"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())
```

## 4.2 Mean reproduction along corr_threshold

Each line is one `n_components` level. The ribbon is the IQR across genes
and cell types. The dashed line marks the best universal `corr_threshold`.

```{r}
#| label: overall-line
#| fig-width: 11
#| fig-height: 5

ggplot(
  overall_grid,
  aes(
    x      = corr_threshold,
    y      = mean_repro,
    colour = factor(n_components),
    fill   = factor(n_components),
    group  = n_components
  )
) +
  geom_ribbon(
    aes(ymin = q25_repro, ymax = q75_repro),
    alpha = 0.12, colour = NA
  ) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  geom_vline(
    xintercept = best_overall$corr_threshold,
    linetype = "dashed", colour = "grey40"
  ) +
  annotate(
    "label",
    x = best_overall$corr_threshold,
    y = best_overall$mean_repro,
    label = glue(
      "best: n_comp={best_overall$n_components}, ",
      "thr={best_overall$corr_threshold}"
    ),
    hjust = -0.05, size = 3.2, label.size = 0
  ) +
  scale_colour_viridis_d(option = "viridis", end = 0.9, name = "n_components") +
  scale_fill_viridis_d  (option = "viridis", end = 0.9, guide = "none") +
  labs(
    title    = "Mean reproduction across `corr_threshold`",
    subtitle = "Ribbon = IQR across (cell_type, gene)",
    x        = "corr_threshold",
    y        = "reproduction (%)"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

## 4.3 Energy and C2ST components

`reproduction_pct` averages `energy_pct` and `c2st_pct`. Plotting each
metric checks that the headline score is not masking opposite behaviour.

```{r}
#| label: overall-line-by-metric
#| fig-width: 11
#| fig-height: 4

param_tunning_long %>%
  select(n_components, corr_threshold, reproduction_pct, energy_pct, c2st_pct) %>%
  pivot_longer(
    cols      = c(reproduction_pct, energy_pct, c2st_pct),
    names_to  = "metric",
    values_to = "pct"
  ) %>%
  group_by(metric, n_components, corr_threshold) %>%
  summarise(mean_pct = mean(pct, na.rm = TRUE), .groups = "drop") %>%
  mutate(metric = factor(
    metric,
    levels = c("reproduction_pct", "energy_pct", "c2st_pct"),
    labels = c("mean (headline)", "energy", "C2ST")
  )) %>%
  ggplot(aes(
    x      = corr_threshold,
    y      = mean_pct,
    colour = factor(n_components),
    group  = n_components
  )) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.5) +
  facet_wrap(~ metric, ncol = 3) +
  scale_colour_viridis_d(option = "viridis", end = 0.9, name = "n_components") +
  labs(
    x = "corr_threshold",
    y = "mean reproduction (%)",
    title = "Each metric's response to (n_components, corr_threshold)"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```


# 5. Agreement across cell types

If the Section 3 default is sensible, each cell type's own optimum should
lie on or near that grid cell. The plots below test that.

## 5.1 Fidelity surface per cell type

```{r}
#| label: per-celltype-heatmap
#| fig-width: 12
#| fig-height: 8

ct_grid <-
  param_tunning_long %>%
  group_by(cell_type, n_components, corr_threshold) %>%
  summarise(
    mean_repro = mean(reproduction_pct, na.rm = TRUE),
    n_obs      = sum(!is.na(reproduction_pct)),
    .groups    = "drop"
  )

best_per_ct <-
  ct_grid %>%
  group_by(cell_type) %>%
  filter(!is.na(mean_repro)) %>%
  slice_max(mean_repro, n = 1, with_ties = FALSE, na_rm = TRUE) %>%
  ungroup()

ct_repro_rng <- range(ct_grid$mean_repro, na.rm = TRUE)

ggplot(
  ct_grid,
  aes(
    x    = factor(corr_threshold),
    y    = factor(n_components),
    fill = mean_repro
  )
) +
  geom_tile(colour = "grey95", linewidth = 0.2) +
  geom_point(
    data  = best_per_ct,
    inherit.aes = FALSE,
    aes(
      x = factor(corr_threshold),
      y = factor(n_components)
    ),
    shape = 21, size = 4, stroke = 0.9,
    fill = NA, colour = "black"
  ) +
  # overall best cell highlighted in every facet for reference
  geom_point(
    data  = best_overall %>% select(n_components, corr_threshold),
    inherit.aes = FALSE,
    aes(
      x = factor(corr_threshold),
      y = factor(n_components)
    ),
    shape = 4, size = 3, stroke = 0.8, colour = "white"
  ) +
  facet_wrap(~ cell_type, ncol = 5) +
  scale_fill_gradient(
    name     = "mean reproduction %",
    low      = "white",
    high     = "#b2182b",
    limits   = ct_repro_rng,
    oob      = scales::squish,
    na.value = "grey85"
  ) +
  labs(
    title    = "Compression-fidelity grid per cell type",
    subtitle = "Black circle = cell type's own best | white cross = overall best",
    x        = "corr_threshold",
    y        = "n_components"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    strip.text  = element_text(size = 9)
  )
```

## 5.2 Best setting per cell type

Tight clustering means cell types agree; spread means a single default
will favour some lineages over others.

```{r}
#| label: best-per-celltype-scatter
#| fig-width: 9
#| fig-height: 5

corr_levels <- sort(unique(param_tunning_long$corr_threshold))
nc_levels   <- sort(unique(param_tunning_long$n_components))

best_per_ct_plot <- best_per_ct %>%
  mutate(
    corr_threshold = factor(corr_threshold, levels = corr_levels),
    n_components   = factor(n_components, levels = nc_levels)
  )

best_overall_plot <- best_overall %>%
  mutate(
    corr_threshold = factor(corr_threshold, levels = corr_levels),
    n_components   = factor(n_components, levels = nc_levels)
  )

ggplot(
  best_per_ct_plot,
  aes(x = corr_threshold, y = n_components,
      colour = cell_type, label = cell_type)
) +
  geom_point(size = 3, alpha = 0.85) +
  ggrepel::geom_text_repel(
    size = 3, max.overlaps = Inf, box.padding = 0.3, show.legend = FALSE
  ) +
  geom_point(
    data = best_overall_plot,
    inherit.aes = FALSE,
    aes(x = corr_threshold, y = n_components),
    shape = 21, size = 6, stroke = 1.2, fill = NA, colour = "black"
  ) +
  scale_colour_viridis_d(option = "turbo", guide = "none") +
  labs(
    title = "Best (n_components, corr_threshold) per cell type",
    subtitle = "Each point sits on a grid value (no jitter) | black circle = overall best",
    x = "corr_threshold", y = "n_components"
  ) +
  theme_bw(base_size = 12)
```

## 5.3 Cost of the universal default

For each cell type, the gap between its own best `reproduction_pct` and
the score at the global best `(n_components, corr_threshold)`.

```{r}
#| label: universal-vs-celltype-best
#| fig-width: 9
#| fig-height: 4

ct_at_universal <-
  ct_grid %>%
  semi_join(best_overall, by = c("n_components", "corr_threshold")) %>%
  select(cell_type, repro_at_universal = mean_repro)

ct_best_gap <-
  best_per_ct %>%
  select(cell_type, repro_at_own_best = mean_repro,
         best_n_components = n_components,
         best_corr_threshold = corr_threshold) %>%
  left_join(ct_at_universal, by = "cell_type") %>%
  mutate(
    gap_pp     = repro_at_own_best - repro_at_universal,
    cell_type  = fct_reorder(cell_type, gap_pp)
  )

ggplot(ct_best_gap, aes(x = cell_type)) +
  geom_segment(
    aes(xend = cell_type,
        y    = repro_at_universal,
        yend = repro_at_own_best),
    colour = "grey60", linewidth = 0.6
  ) +
  geom_point(aes(y = repro_at_universal, colour = "universal setting"),
             size = 3) +
  geom_point(aes(y = repro_at_own_best,  colour = "own best"),
             size = 3) +
  geom_text(
    aes(y = repro_at_own_best,
        label = sprintf("+%.2f pp", gap_pp)),
    nudge_y = 0.05, size = 3, hjust = 0
  ) +
  scale_colour_manual(
    name   = NULL,
    values = c("universal setting" = "#2c7fb8",
               "own best"          = "#b2182b")
  ) +
  coord_flip() +
  labs(
    title    = "Cost of using one universal (n_components, corr_threshold)",
    subtitle = glue(
      "Universal = (n_components={best_overall$n_components}, ",
      "corr_threshold={best_overall$corr_threshold})"
    ),
    x = NULL,
    y = "mean reproduction (%)"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

## 5.4 Rank agreement between cell types

Spearman correlation of mean reproduction ranks across grid cells. Values
near 1 mean cell types order parameters similarly; low values mean
genuinely different preferences.

```{r}
#| label: celltype-rank-agreement
#| fig-width: 7
#| fig-height: 6

ct_rank_wide <- ct_grid %>%
  unite("grid_cell", n_components, corr_threshold, sep = "_") %>%
  select(cell_type, grid_cell, mean_repro) %>%
  pivot_wider(names_from = cell_type, values_from = mean_repro)

ct_rank_cor <- cor(
  ct_rank_wide %>% select(-grid_cell),
  method = "spearman",
  use = "pairwise.complete.obs"
)

ct_rank_cor_df <- as.data.frame(as.table(ct_rank_cor), stringsAsFactors = FALSE)
names(ct_rank_cor_df) <- c("ct_x", "ct_y", "rho")
ct_rank_cor_df <- ct_rank_cor_df %>%
  mutate(
    ct_x = factor(ct_x, levels = colnames(ct_rank_cor)),
    ct_y = factor(ct_y, levels = colnames(ct_rank_cor))
  )

ggplot(ct_rank_cor_df, aes(ct_x, ct_y, fill = rho)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", rho)), size = 2.8) +
  scale_fill_gradient2(
    name     = "Spearman rho",
    low      = "#2c7fb8", mid = "white", high = "#b2182b",
    midpoint = 0, limits = c(-1, 1)
  ) +
  labs(
    title    = "Across-cell-type agreement on grid-cell ranking",
    subtitle = "Spearman correlation over 63 (n_components, corr_threshold) cells",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid  = element_blank()
  )
```


# 6. Gene abundance and parameter choice

For each gene we take the grid cell with highest `reproduction_pct` and
ask whether that optimum shifts with pseudobulk `mean_log2_cpm`. We also
plot fidelity at the Section 3 universal setting across abundance—useful
for spotting genes that need per-gene tuning.

```{r}
#| label: gene-best-vs-abundance

gene_best <-
  param_tunning_long %>%
  filter(!is.na(reproduction_pct)) %>%
  group_by(.feature, cell_type) %>%
  slice_max(reproduction_pct, n = 1, with_ties = FALSE, na_rm = TRUE) %>%
  ungroup() %>%
  filter(!is.na(mean_log2_cpm)) %>%
  # carry the per-gene abundance metrics (constant within .feature/cell_type)
  mutate(
    abundance_bin = cut(
      mean_log2_cpm,
      breaks = quantile(mean_log2_cpm, c(0, 1/3, 2/3, 1), na.rm = TRUE),
      include.lowest = TRUE,
      labels = c("low", "mid", "high")
    )
  )

cat("Per-gene best parameters and abundance bins:\n")
print(
  gene_best %>%
    count(abundance_bin, n_components, corr_threshold) %>%
    arrange(desc(n))
)
```

## 6.1 Best parameters vs expression

```{r}
#| label: best-vs-abundance
#| fig-width: 11
#| fig-height: 4.5

p_nc <-
  ggplot(gene_best, aes(x = mean_log2_cpm, y = n_components)) +
  geom_jitter(aes(colour = cell_type),
              height = 0.15, width = 0, size = 2, alpha = 0.85) +
  geom_smooth(method = "lm", se = TRUE, colour = "grey20",
              linetype = "dashed") +
  scale_y_continuous(breaks = sort(unique(param_tunning_long$n_components))) +
  scale_colour_viridis_d(option = "turbo", guide = "none") +
  labs(
    x = "gene mean log2 CPM (pseudobulk)",
    y = "best n_components",
    title = "Best n_components vs gene abundance"
  ) +
  theme_bw(base_size = 12)

p_ct <-
  ggplot(gene_best, aes(x = mean_log2_cpm, y = corr_threshold)) +
  geom_jitter(aes(colour = cell_type),
              height = 0.005, width = 0, size = 2, alpha = 0.85) +
  geom_smooth(method = "lm", se = TRUE, colour = "grey20",
              linetype = "dashed") +
  scale_y_continuous(
    breaks = sort(unique(param_tunning_long$corr_threshold))
  ) +
  scale_colour_viridis_d(option = "turbo", name = "cell_type") +
  labs(
    x = "gene mean log2 CPM (pseudobulk)",
    y = "best corr_threshold",
    title = "Best corr_threshold vs gene abundance"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "right", legend.text = element_text(size = 7),
        legend.key.size = unit(0.4, "cm"))

p_nc + p_ct + plot_layout(widths = c(1, 1.25))
```

```{r}
#| label: abundance-tests

cor_test_nc <- with(
  gene_best,
  suppressWarnings(cor.test(mean_log2_cpm, n_components, method = "spearman"))
)
cor_test_ct <- with(
  gene_best,
  suppressWarnings(cor.test(mean_log2_cpm, corr_threshold, method = "spearman"))
)

tibble(
  parameter    = c("n_components", "corr_threshold"),
  spearman_rho = c(cor_test_nc$estimate, cor_test_ct$estimate),
  p_value      = c(cor_test_nc$p.value, cor_test_ct$p.value)
)
```

## 6.2 Universal setting vs abundance

A flat trend supports one default for all genes; a slope at low abundance
flags genes that may need separate settings.

```{r}
#| label: universal-vs-abundance
#| fig-width: 9
#| fig-height: 5

param_tunning_long %>%
  semi_join(best_overall, by = c("n_components", "corr_threshold")) %>%
  filter(!is.na(reproduction_pct), !is.na(mean_log2_cpm)) %>%
  ggplot(aes(x = mean_log2_cpm, y = reproduction_pct)) +
  geom_point(aes(colour = cell_type), size = 2, alpha = 0.85) +
  geom_smooth(method = "loess", se = TRUE, colour = "grey20") +
  scale_colour_viridis_d(option = "turbo", name = "cell_type") +
  labs(
    title    = "Universal setting performance vs gene abundance",
    subtitle = glue(
      "At n_components={best_overall$n_components}, ",
      "corr_threshold={best_overall$corr_threshold}"
    ),
    x = "gene mean log2 CPM (pseudobulk)",
    y = "reproduction (%)"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "right",
        legend.text = element_text(size = 7),
        legend.key.size = unit(0.4, "cm"))
```

## 6.3 Grid surfaces by abundance tercile

Different panels suggest abundance-specific optima; similar panels support
a single universal default.

```{r}
#| label: grid-by-abundance
#| fig-width: 11
#| fig-height: 4.5

abundance_keyed <-
  param_tunning_long %>%
  distinct(.feature, cell_type, mean_log2_cpm) %>%
  filter(!is.na(mean_log2_cpm)) %>%
  mutate(
    abundance_bin = cut(
      mean_log2_cpm,
      breaks = quantile(mean_log2_cpm, c(0, 1/3, 2/3, 1), na.rm = TRUE),
      include.lowest = TRUE,
      labels = c("low", "mid", "high")
    )
  ) %>%
  select(.feature, cell_type, abundance_bin)

grid_by_bin <-
  param_tunning_long %>%
  inner_join(abundance_keyed, by = c(".feature", "cell_type")) %>%
  group_by(abundance_bin, n_components, corr_threshold) %>%
  summarise(
    mean_repro = mean(reproduction_pct, na.rm = TRUE),
    .groups    = "drop"
  )

best_per_bin <-
  grid_by_bin %>%
  group_by(abundance_bin) %>%
  filter(!is.na(mean_repro), !is.na(abundance_bin)) %>%
  slice_max(mean_repro, n = 1, with_ties = FALSE, na_rm = TRUE) %>%
  ungroup()

ggplot(
  grid_by_bin,
  aes(
    x    = factor(corr_threshold),
    y    = factor(n_components),
    fill = mean_repro
  )
) +
  geom_tile(colour = "grey95", linewidth = 0.2) +
  geom_text(aes(label = sprintf("%.1f", mean_repro)),
            colour = "black", size = 2.6) +
  geom_point(
    data = best_per_bin,
    inherit.aes = FALSE,
    aes(
      x = factor(corr_threshold),
      y = factor(n_components)
    ),
    shape = 21, size = 6, stroke = 1.1,
    fill = NA, colour = "black"
  ) +
  facet_wrap(~ abundance_bin, nrow = 1,
             labeller = label_both) +
  scale_fill_gradient(
    name     = "mean reproduction %",
    low      = "white",
    high     = "#b2182b",
    na.value = "grey85"
  ) +
  labs(
    title    = "Grid surface stratified by gene-abundance tercile",
    subtitle = "Circled cell = best (n_components, corr_threshold) within each tercile",
    x        = "corr_threshold",
    y        = "n_components"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
  )
```


# 7. Summary

Recommended defaults from the grid search:

```{r}
#| label: summary-table

bind_rows(
  best_overall %>%
    transmute(
      scope = "overall (universal)",
      n_components, corr_threshold,
      mean_reproduction_pct = mean_repro
    ),
  best_per_ct %>%
    transmute(
      scope = paste0("cell_type: ", cell_type),
      n_components, corr_threshold,
      mean_reproduction_pct = mean_repro
    ),
  best_per_bin %>%
    transmute(
      scope = paste0("abundance: ", abundance_bin),
      n_components, corr_threshold,
      mean_reproduction_pct = mean_repro
    )
)
```

```{r}
sessionInfo()
```

